---
title: "Bayesplot and the Bayesian workflow"
author: "bayesplot team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Bayesplot and the Bayesian workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/SETTINGS-knitr.txt"}
```
```{r pkgs, include=FALSE}
library("ggplot2")
library("rstanarm")
library("bayesplot")
bayesplot_theme_set()
set.seed(1234)
```

## Introduction

A Bayesian analysis is more than fitting a single model and reading off the
estimates. The _Bayesian workflow_ (Gelman et al., 2020; Gabry et al., 2019)
involves iterating through several stages:

1. **Prior predictive checking** -- do the priors generate plausible data?
2. **Fitting the model** -- obtain posterior draws via MCMC.
3. **MCMC diagnostics** -- did the sampler explore the posterior adequately?
4. **Posterior summarization** -- what did we learn about the parameters?
5. **Posterior predictive checking** -- does the fitted model reproduce the data?
6. **Model comparison / revision** -- can we do better?

This vignette walks through each stage with a single running example, showing
which **bayesplot** functions are useful at every step.

### Setup

```{r setup, eval=FALSE}
library("bayesplot")
library("ggplot2")
library("rstanarm")
```

### Running example

We use the `kidiq` dataset bundled with **rstanarm**, which contains IQ scores
of children and mothers along with whether the mother completed high school.

```{r data}
data("kidiq", package = "rstanarm")
head(kidiq)
```

Our goal is to model the child's IQ score (`kid_score`) as a function of the
mother's IQ (`mom_iq`) and high-school completion status (`mom_hs`).

## Step 1: Prior predictive checking

Before looking at the data, it is useful to check whether our priors produce
sensible predictions. We fit the model with `prior_PD = TRUE` so that the
likelihood is ignored and we sample only from the prior predictive distribution.

```{r prior-pred, message=FALSE}
fit_prior <- stan_glm(
  kid_score ~ mom_hs + mom_iq,
  data = kidiq,
  prior_PD = TRUE,
  seed = 111,
  refresh = 0
)
```

We then draw from the prior predictive distribution and visualise it with
`ppd_dens_overlay`:

```{r ppd-prior}
yrep_prior <- posterior_predict(fit_prior, draws = 100)
ppd_dens_overlay(ypred = yrep_prior[1:50, ])
```

If the prior predictive densities cover an implausibly wide range (e.g.
negative IQ scores, or values in the thousands), it is a signal to tighten the
priors. With **rstanarm**'s default weakly informative priors the range is
already reasonable for this problem, so we proceed.

## Step 2: Fit the model

```{r fit, message=FALSE}
fit <- stan_glm(
  kid_score ~ mom_hs + mom_iq,
  data = kidiq,
  seed = 111,
  refresh = 0
)
print(fit)
```

## Step 3: MCMC diagnostics

### Trace plots

A quick visual check that the chains mixed well:

```{r trace}
posterior <- as.array(fit)
color_scheme_set("mix-blue-red")
mcmc_trace(posterior, pars = c("mom_hs", "mom_iq", "sigma"))
```

All chains should be exploring the same region, with no sustained trends or
sticky patches.

### R-hat

```{r rhat}
color_scheme_set("brightblue")
mcmc_rhat(rhat(fit)) + yaxis_text(hjust = 1)
```

All $\hat{R}$ values should be close to 1.0 (below 1.01 is ideal).

### Effective sample size

```{r neff}
mcmc_neff(neff_ratio(fit)) + yaxis_text(hjust = 1)
```

Ratios well above 0.1 indicate low autocorrelation and efficient sampling.

### Autocorrelation

```{r acf}
mcmc_acf(posterior, pars = c("mom_hs", "mom_iq"), lags = 15)
```

Autocorrelation should drop to near zero within a few lags.

## Step 4: Posterior summarization

### Uncertainty intervals

```{r intervals}
color_scheme_set("red")
mcmc_intervals(posterior, pars = c("mom_hs", "mom_iq", "sigma"))
```

### Posterior densities

```{r areas}
mcmc_areas(posterior, pars = c("mom_hs", "mom_iq"), prob = 0.8)
```

Both `mom_hs` and `mom_iq` have posterior mass clearly away from zero,
indicating positive associations with the child's IQ.

### Bivariate relationships

```{r scatter}
color_scheme_set("gray")
mcmc_scatter(posterior, pars = c("mom_hs", "mom_iq"), size = 1, alpha = 0.3)
```

## Step 5: Posterior predictive checks

Now we check whether the model can reproduce the observed data.

```{r ppc-data}
y <- kidiq$kid_score
yrep <- posterior_predict(fit, draws = 500)
```

### Density overlay

```{r ppc-dens}
color_scheme_set("brightblue")
ppc_dens_overlay(y, yrep[1:50, ])
```

The replicated densities should closely track the observed density.

### Test statistics

We can check specific summaries. For instance, does the model replicate the
standard deviation of the data?

```{r ppc-stat, message=FALSE}
ppc_stat(y, yrep, stat = "sd")
```

And the proportion of scores above 100:

```{r ppc-stat-custom, message=FALSE}
ppc_stat(y, yrep, stat = function(x) mean(x > 100))
```

### Intervals per observation

```{r ppc-intervals}
ppc_intervals(y, yrep, x = kidiq$mom_iq, prob = 0.5, prob_outer = 0.9) +
  labs(x = "Mother's IQ", y = "Child's IQ")
```

### Grouped checks

```{r ppc-grouped, message=FALSE}
ppc_stat_grouped(y, yrep, group = kidiq$mom_hs, stat = "median")
```

## Step 6: Model comparison and iteration

If the checks above reveal systematic problems -- for example, if the model
under-predicts for children of high-IQ mothers -- we might revise the model
(add an interaction, use a different likelihood, etc.) and repeat the cycle.

Model comparison tools such as **loo** can be used alongside **bayesplot**'s
LOO-PIT plots to evaluate relative and absolute model fit:

```{r loo-pit, eval=FALSE}
library("loo")
loo1 <- loo(fit, save_psis = TRUE)
ppc_loo_pit_overlay(y, yrep, lw = weights(loo1$psis_object))
```

## Summary

The table below maps each workflow stage to the most useful **bayesplot**
functions:

| Stage | Key functions |
|---|---|
| Prior predictive check | `ppd_dens_overlay()`, `ppd_hist()`, `ppd_stat()` |
| MCMC diagnostics | `mcmc_trace()`, `mcmc_rhat()`, `mcmc_neff()`, `mcmc_acf()`, `mcmc_nuts_divergence()`, `mcmc_nuts_energy()` |
| Posterior summaries | `mcmc_intervals()`, `mcmc_areas()`, `mcmc_hist()`, `mcmc_dens()`, `mcmc_pairs()` |
| Posterior predictive checks | `ppc_dens_overlay()`, `ppc_stat()`, `ppc_intervals()`, `ppc_error_hist()`, `ppc_bars()` |
| Model comparison (LOO) | `ppc_loo_pit_overlay()`, `ppc_loo_pit_qq()`, `ppc_loo_intervals()` |

For a complete lookup table, see the
[_Which plot should I use?_](https://mc-stan.org/bayesplot/articles/which-plot.html)
vignette.

## References

Gabry, J., Simpson, D., Vehtari, A., Betancourt, M. and Gelman, A. (2019),
Visualization in Bayesian workflow. _J. R. Stat. Soc. A_, 182: 389-402.
\doi{10.1111/rssa.12378}

Gelman, A., Vehtari, A., Simpson, D., et al. (2020), Bayesian Workflow.
https://arxiv.org/abs/2011.01808
