---
title: "Customizing bayesplot outputs"
author: "bayesplot team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Customizing bayesplot outputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/SETTINGS-knitr.txt"}
```
```{r pkgs, include=FALSE}
library("ggplot2")
library("bayesplot")
bayesplot_theme_set()
```

## Introduction

Every plot returned by **bayesplot** is a standard **ggplot2** object. That
means you can modify titles, labels, scales, layers, facets, and themes using
the same `+` operator you already know. This vignette shows you how.

### Setup

```{r setup, eval=FALSE}
library("bayesplot")
library("ggplot2")
```

We'll use the built-in example data throughout:

```{r example-data}
x <- example_mcmc_draws()
dim(x)
y <- example_y_data()
yrep <- example_yrep_draws()
```

## Adding layers and labels with `+`

Because every bayesplot is a ggplot, you can add geoms, labels, and
coordinate adjustments in the usual way:

```{r add-layers, message=FALSE}
mcmc_hist(x, pars = "beta[1]") +
  labs(
    title = "Posterior distribution of beta[1]",
    subtitle = "with a vertical line at zero"
  ) +
  vline_0(linetype = 2, color = "gray50")
```

The `vline_0()` helper above is one of several convenience functions provided
by **bayesplot** (see [Convenience helpers](#convenience-helpers) below).

### Restricting axis ranges

```{r xlim, message=FALSE, warning=FALSE}
ppc_dens_overlay(y, yrep[1:30, ]) +
  xlim(-5, 60) +
  labs(x = "Outcome", y = "Density")
```

### Annotating specific values

```{r annotate, message=FALSE}
mcmc_areas(x, pars = c("beta[1]", "beta[2]"), prob = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red")
```

## Color schemes

**bayesplot** includes a flexible system for changing the color palette
globally. Call `color_scheme_set()` with the name of a preset scheme:

```{r color-scheme-set}
color_scheme_set("red")
mcmc_intervals(x, pars = c("beta[1]", "beta[2]", "sigma"))
```

```{r color-scheme-teal}
color_scheme_set("teal")
mcmc_intervals(x, pars = c("beta[1]", "beta[2]", "sigma"))
```

### Available preset schemes

The built-in schemes include:

`"blue"`, `"brightblue"`, `"gray"`, `"darkgray"`, `"green"`, `"pink"`,
`"purple"`, `"red"`, `"teal"`, `"yellow"`, and the viridis family:
`"viridis"`, `"viridisA"` through `"viridisE"`.

You can also mix two schemes (e.g. `"mix-blue-red"`) or use ColorBrewer
palettes (e.g. `"brewer-PuBuGn"`):

```{r color-scheme-mix}
color_scheme_set("mix-blue-red")
mcmc_trace(x, pars = "sigma")
```

### Viewing and comparing schemes

```{r color-scheme-view, fig.width=7}
color_scheme_view(c("blue", "red", "green", "teal"))
```

### Custom color schemes

Pass a character vector of exactly six colors (lightest to darkest) for a
fully custom palette:

```{r custom-scheme}
color_scheme_set(c("#F0F0F0", "#C0C0C0", "#909090", "#606060", "#303030", "#000000"))
mcmc_intervals(x, pars = c("beta[1]", "beta[2]", "sigma"))
```

```{r reset-scheme, include=FALSE}
color_scheme_set("blue")
```

## Themes

### The default theme

**bayesplot** uses its own `theme_default()` based on `ggplot2::theme_bw()`.
You can customise the base font size and family:

```{r theme-default}
bayesplot_theme_set(theme_default(base_size = 14, base_family = "sans"))
mcmc_hist(x, pars = "sigma")
```

### Using any ggplot2 theme

You can switch to any ggplot2 theme:

```{r theme-minimal}
bayesplot_theme_set(ggplot2::theme_minimal())
mcmc_hist(x, pars = "sigma")
```

```{r theme-reset, include=FALSE}
bayesplot_theme_set()
```

### Updating individual theme elements

Use `bayesplot_theme_update()` to modify specific elements without replacing
the entire theme:

```{r theme-update}
bayesplot_theme_update(
  plot.title = element_text(face = "bold", size = 16),
  axis.title = element_text(size = 12)
)
mcmc_areas(x, pars = c("beta[1]", "beta[2]")) +
  ggtitle("Updated theme elements")
```

```{r theme-reset-2, include=FALSE}
bayesplot_theme_set()
```

### Per-plot theme overrides

You can always add a `theme()` call to a single plot without affecting the
global setting:

```{r per-plot-theme}
mcmc_dens(x, pars = c("beta[1]", "beta[2]")) +
  theme(
    panel.background = element_rect(fill = "gray95"),
    legend.position = "bottom"
  )
```

## Convenience helpers {#convenience-helpers}

**bayesplot** exports several small helper functions that make common
modifications easier:

### Reference lines

| Function | Description |
|---|---|
| `vline_at(v, ...)` | Vertical line(s) at `v` (or `fun(v)`) |
| `hline_at(v, ...)` | Horizontal line(s) at `v` (or `fun(v)`) |
| `vline_0(...)` | Vertical line at zero |
| `hline_0(...)` | Horizontal line at zero |
| `abline_01(...)` | Diagonal identity line (slope 1, intercept 0) |

```{r helpers-lines}
mcmc_intervals(x, pars = c("beta[1]", "beta[2]")) +
  vline_0(linetype = "dotted", linewidth = 0.5)
```

### Axis and legend control

| Function | Description |
|---|---|
| `xaxis_text(...)` | Modify or toggle x-axis tick labels |
| `yaxis_text(...)` | Modify or toggle y-axis tick labels |
| `xaxis_ticks(...)` | Modify x-axis tick marks |
| `yaxis_ticks(...)` | Modify y-axis tick marks |
| `xaxis_title(...)` | Modify or toggle x-axis title |
| `yaxis_title(...)` | Modify or toggle y-axis title |
| `legend_move(position)` | Move the legend (e.g., `"bottom"`) |
| `legend_none()` | Remove the legend |
| `legend_text(...)` | Modify legend text |

```{r helpers-axis}
mcmc_intervals(x) +
  yaxis_text(hjust = 1, size = 10, face = "italic") +
  xaxis_title(size = 12, face = "bold")
```

### Background control

| Function | Description |
|---|---|
| `plot_bg(...)` | Modify the plot background |
| `panel_bg(...)` | Modify the panel background |
| `facet_bg(...)` | Modify the facet strip background |
| `facet_text(...)` | Modify the facet strip text |
| `grid_lines(...)` | Add major grid lines |

```{r helpers-bg}
color_scheme_set("purple")
mcmc_hist(x, pars = "beta[1]") +
  panel_bg(fill = "gray95") +
  grid_lines(color = "white", linewidth = 0.3)
```

### Function overlays

`overlay_function()` is a wrapper around `ggplot2::stat_function()` for
overlaying a mathematical function onto a plot:

```{r overlay}
color_scheme_set("brightblue")
mcmc_hist(x, pars = "sigma", freq = FALSE) +
  overlay_function(fun = dnorm, args = list(mean = 1, sd = 0.2), color = "red")
```

## Using `*_data()` functions for fully custom plots

Many bayesplot functions have a `*_data()` companion that returns the
pre-processed data as a tidy data frame. This gives you complete control over
the final ggplot:

```{r data-custom}
d <- mcmc_intervals_data(x, pars = c("beta[1]", "beta[2]", "sigma"))
d

ggplot(d, aes(y = parameter, x = m)) +
  geom_point(size = 3) +
  geom_linerange(aes(xmin = ll, xmax = hh), linewidth = 0.5) +
  geom_linerange(aes(xmin = l, xmax = h), linewidth = 1.2) +
  labs(x = "Posterior estimate", y = NULL) +
  theme_minimal()
```

See `available_mcmc(plots_only = FALSE)` and
`available_ppc(plots_only = FALSE)` for a full list of data functions.

## Composing multiple plots with `bayesplot_grid`

`bayesplot_grid()` arranges several bayesplot objects into a grid:

```{r grid, fig.width=7, fig.height=4}
color_scheme_set("blue")
p1 <- mcmc_hist(x, pars = "beta[1]")
p2 <- mcmc_hist(x, pars = "beta[2]")
bayesplot_grid(p1, p2, grid_args = list(ncol = 2),
               subtitles = c("Beta 1", "Beta 2"))
```

You can also use **patchwork** or **cowplot** as you would with any ggplot
objects.

## Summary

| Want to... | How |
|---|---|
| Change title/labels | `+ labs(...)` |
| Restrict axis range | `+ xlim(...)` / `+ coord_cartesian(...)` |
| Change color palette | `color_scheme_set("name")` |
| Change theme globally | `bayesplot_theme_set(theme)` |
| Tweak one theme element | `bayesplot_theme_update(...)` |
| Add reference lines | `+ vline_at(...)`, `+ hline_0()`, etc. |
| Toggle axis labels | `+ xaxis_text(FALSE)`, `+ yaxis_text(hjust = 1)` |
| Build fully custom plot | Use the `*_data()` companion function |
| Arrange plots in grid | `bayesplot_grid(...)` |

## Further reading

* [`?bayesplot-colors`](https://mc-stan.org/bayesplot/reference/bayesplot-colors.html)
  -- full documentation for color scheme functions
* [`?bayesplot-helpers`](https://mc-stan.org/bayesplot/reference/bayesplot-helpers.html)
  -- full documentation for convenience helpers
* [`?theme_default`](https://mc-stan.org/bayesplot/reference/theme_default.html)
  -- default theme and theme management
