---
title: "Example Gallery"
author: "bayesplot team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{Example gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
stopifnot(require("knitr"))
library("bayesplot")
library("ggplot2")

knitr::opts_chunk$set(
  dev = "png",
  dpi = 96,
  fig.width = 3.5,
  fig.height = 2.8,
  out.width = "100%",
  fig.align = "center",
  comment = NA,
  eval = if (isTRUE(exists("params"))) params$EVAL else FALSE,
  message = FALSE,
  warning = FALSE
)

bayesplot_theme_set(theme_default(base_size = 8))
color_scheme_set("brightblue")
set.seed(8420)

# Prepare shared example data
y <- example_y_data()
yrep <- example_yrep_draws()
x <- example_x_data()
group <- example_group_data()
draws <- example_mcmc_draws(params = 6)
```

```{r gallery-helpers, echo=FALSE}
# Base URL for reference pages
ref_base <- "https://mc-stan.org/bayesplot/reference/"

# Helper: emit one gallery card (clickable thumbnail + linked label)
gallery_card <- function(chunk_label, ref_page, fn_name) {
  url <- paste0(ref_base, ref_page)
  cat('<div class="gallery-card">\n')
  cat(paste0('<a href="', url, '">'))
  cat(paste0('<img src="', fig_chunk(chunk_label, 'png'),
             '" alt="', fn_name, '()" loading="lazy" />\n'))
  cat('</a>\n')
  cat(paste0('<div class="card-label"><a href="', url, '">',
             fn_name, '()</a></div>\n'))
  cat('</div>\n')
}

# Helper for conditional cards (e.g. requiring optional packages)
gallery_card_cond <- function(chunk_label, ref_page, fn_name, condition,
                              fallback = "Requires survival package") {
  url <- paste0(ref_base, ref_page)
  cat('<div class="gallery-card">\n')
  if (condition) {
    cat(paste0('<a href="', url, '">'))
    cat(paste0('<img src="', fig_chunk(chunk_label, 'png'),
               '" alt="', fn_name, '()" loading="lazy" />\n'))
    cat('</a>\n')
  } else {
    cat(paste0('<div style="padding:40px;text-align:center;color:#999;">',
               fallback, '</div>\n'))
  }
  cat(paste0('<div class="card-label"><a href="', url, '">',
             fn_name, '()</a></div>\n'))
  cat('</div>\n')
}
```

```{css gallery-styles, echo=FALSE}
/* ---- Gallery grid layout ---- */
.gallery-grid {
  display: grid;
  grid-template-columns: repeat(auto-fill, minmax(260px, 1fr));
  gap: 1.2em;
  margin-bottom: 2em;
}
.gallery-card {
  border: 1px solid #ddd;
  border-radius: 6px;
  overflow: hidden;
  background: #fff;
  transition: box-shadow 0.2s;
}
.gallery-card:hover {
  box-shadow: 0 2px 12px rgba(0,0,0,0.12);
}
.gallery-card img {
  width: 100%;
  display: block;
}
.gallery-card .card-label {
  padding: 0.5em 0.75em;
  font-family: 'SFMono-Regular', 'Menlo', monospace;
  font-size: 0.85em;
  text-align: center;
  border-top: 1px solid #eee;
}
.gallery-card .card-label a {
  color: #2171b5;
  text-decoration: none;
}
.gallery-card .card-label a:hover {
  text-decoration: underline;
}

/* make the vignette wider for the gallery layout */
.main-container {
  max-width: 960px !important;
}
```

This gallery provides a visual overview of all the plotting functions available
in the **bayesplot** package, organized by category. Click on any function name
to see its full documentation.

> **See also:** The
> [Plotting MCMC draws](plotting-mcmc-draws.html),
> [Visual MCMC diagnostics](visual-mcmc-diagnostics.html), and
> [Graphical PPCs](graphical-ppcs.html) vignettes provide in-depth tutorials.


# PPC: Posterior Predictive Checks {#ppc}

Plotting functions for graphical posterior predictive checks, comparing
observed data `y` to simulated data `yrep` from the posterior predictive
distribution. See `help("PPC-overview")` for details.

## Distributions {#ppc-distributions}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_hist, echo=FALSE, fig.show='hide'}
p <- ppc_hist(y, yrep[1:5, ])
print(p)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_hist", "PPC-distributions.html", "ppc_hist")
```

```{r ppc_dens, echo=FALSE, fig.show='hide'}
ppc_dens(y, yrep[200:202, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_dens", "PPC-distributions.html", "ppc_dens")
```

```{r ppc_dens_overlay, echo=FALSE, fig.show='hide'}
ppc_dens_overlay(y, yrep[1:25, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_dens_overlay", "PPC-distributions.html", "ppc_dens_overlay")
```

```{r ppc_ecdf_overlay, echo=FALSE, fig.show='hide'}
ppc_ecdf_overlay(y, yrep[sample(nrow(yrep), 25), ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_ecdf_overlay", "PPC-distributions.html", "ppc_ecdf_overlay")
```

```{r ppc_freqpoly, echo=FALSE, fig.show='hide'}
ppc_freqpoly(y, yrep[1:3, ], alpha = 0.1, size = 1)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_freqpoly", "PPC-distributions.html", "ppc_freqpoly")
```

```{r ppc_freqpoly_grouped, echo=FALSE, fig.show='hide'}
ppc_freqpoly_grouped(y, yrep[1:3, ], group, freq = FALSE)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_freqpoly_grouped", "PPC-distributions.html", "ppc_freqpoly_grouped")
```

```{r ppc_boxplot, echo=FALSE, fig.show='hide'}
ppc_boxplot(y, yrep[1:8, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_boxplot", "PPC-distributions.html", "ppc_boxplot")
```

```{r ppc_violin_grouped, echo=FALSE, fig.show='hide'}
ppc_violin_grouped(y, yrep, group, size = 1.5)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_violin_grouped", "PPC-distributions.html", "ppc_violin_grouped")
```

```{r ppc_dens_overlay_grouped, echo=FALSE, fig.show='hide'}
ppc_dens_overlay_grouped(y, yrep[1:25, ], group = group)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_dens_overlay_grouped", "PPC-distributions.html", "ppc_dens_overlay_grouped")
```

```{r ppc_ecdf_overlay_grouped, echo=FALSE, fig.show='hide'}
ppc_ecdf_overlay_grouped(y, yrep[1:25, ], group = group)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_ecdf_overlay_grouped", "PPC-distributions.html", "ppc_ecdf_overlay_grouped")
```

```{r ppc_pit_ecdf, echo=FALSE, fig.show='hide'}
ppc_pit_ecdf(y, yrep, prob = 0.99, plot_diff = FALSE)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_pit_ecdf", "PPC-distributions.html", "ppc_pit_ecdf")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Test Statistics {#ppc-test-statistics}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_stat, echo=FALSE, fig.show='hide'}
ppc_stat(y, yrep, stat = "mean")
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_stat", "PPC-test-statistics.html", "ppc_stat")
```

```{r ppc_stat_2d, echo=FALSE, fig.show='hide'}
ppc_stat_2d(y, yrep, stat = c("mean", "sd"))
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_stat_2d", "PPC-test-statistics.html", "ppc_stat_2d")
```

```{r ppc_stat_grouped, echo=FALSE, fig.show='hide'}
ppc_stat_grouped(y, yrep, group = group, stat = "median")
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_stat_grouped", "PPC-test-statistics.html", "ppc_stat_grouped")
```

```{r ppc_stat_freqpoly, echo=FALSE, fig.show='hide'}
ppc_stat_freqpoly(y, yrep, stat = "mean")
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_stat_freqpoly", "PPC-test-statistics.html", "ppc_stat_freqpoly")
```

```{r ppc_stat_freqpoly_grouped, echo=FALSE, fig.show='hide'}
ppc_stat_freqpoly_grouped(y, yrep, group = group, stat = "median")
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_stat_freqpoly_grouped", "PPC-test-statistics.html", "ppc_stat_freqpoly_grouped")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Intervals {#ppc-intervals}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_intervals, echo=FALSE, fig.show='hide'}
ppc_intervals(y, yrep, x = x)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_intervals", "PPC-intervals.html", "ppc_intervals")
```

```{r ppc_ribbon, echo=FALSE, fig.show='hide'}
ppc_ribbon(y, yrep, x = x)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_ribbon", "PPC-intervals.html", "ppc_ribbon")
```

```{r ppc_intervals_grouped, echo=FALSE, fig.show='hide'}
ppc_intervals_grouped(y, yrep, x = x, group = group)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_intervals_grouped", "PPC-intervals.html", "ppc_intervals_grouped")
```

```{r ppc_ribbon_grouped, echo=FALSE, fig.show='hide'}
ppc_ribbon_grouped(y, yrep, x = x, group = group)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_ribbon_grouped", "PPC-intervals.html", "ppc_ribbon_grouped")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Scatterplots {#ppc-scatterplots}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_scatter, echo=FALSE, fig.show='hide'}
ppc_scatter(y, yrep[1:3, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_scatter", "PPC-scatterplots.html", "ppc_scatter")
```

```{r ppc_scatter_avg, echo=FALSE, fig.show='hide'}
ppc_scatter_avg(y, yrep)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_scatter_avg", "PPC-scatterplots.html", "ppc_scatter_avg")
```

```{r ppc_scatter_avg_grouped, echo=FALSE, fig.show='hide'}
ppc_scatter_avg_grouped(y, yrep, group = group)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_scatter_avg_grouped", "PPC-scatterplots.html", "ppc_scatter_avg_grouped")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Errors {#ppc-errors}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_error_hist, echo=FALSE, fig.show='hide'}
ppc_error_hist(y, yrep[1:3, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_error_hist", "PPC-errors.html", "ppc_error_hist")
```

```{r ppc_error_hist_grouped, echo=FALSE, fig.show='hide'}
ppc_error_hist_grouped(y, yrep[1:3, ], group = group)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_error_hist_grouped", "PPC-errors.html", "ppc_error_hist_grouped")
```

```{r ppc_error_scatter, echo=FALSE, fig.show='hide'}
ppc_error_scatter(y, yrep[1:3, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_error_scatter", "PPC-errors.html", "ppc_error_scatter")
```

```{r ppc_error_scatter_avg, echo=FALSE, fig.show='hide'}
ppc_error_scatter_avg(y, yrep)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_error_scatter_avg", "PPC-errors.html", "ppc_error_scatter_avg")
```

```{r ppc_error_scatter_avg_vs_x, echo=FALSE, fig.show='hide'}
ppc_error_scatter_avg_vs_x(y, yrep, x = x)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_error_scatter_avg_vs_x", "PPC-errors.html", "ppc_error_scatter_avg_vs_x")
```

```{r ppc_error_binned, echo=FALSE, fig.show='hide'}
ppc_error_binned(y, yrep[1:3, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_error_binned", "PPC-errors.html", "ppc_error_binned")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Discrete {#ppc-discrete}

```{r setup-discrete, echo=FALSE}
# Create simple discrete data for these plots
set.seed(123)
y_disc <- rpois(200, lambda = 3)
yrep_disc <- matrix(rpois(200 * 100, lambda = 3), nrow = 100)
group_disc <- factor(rep(c("A", "B"), each = 100))
```

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_bars, echo=FALSE, fig.show='hide'}
ppc_bars(y_disc, yrep_disc[1:50, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_bars", "PPC-discrete.html", "ppc_bars")
```

```{r ppc_bars_grouped, echo=FALSE, fig.show='hide'}
ppc_bars_grouped(y_disc, yrep_disc[1:50, ], group = group_disc)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_bars_grouped", "PPC-discrete.html", "ppc_bars_grouped")
```

```{r ppc_rootogram, echo=FALSE, fig.show='hide'}
ppc_rootogram(y_disc, yrep_disc)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_rootogram", "PPC-discrete.html", "ppc_rootogram")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## LOO Predictive Checks {#ppc-loo}

```{r setup-loo, echo=FALSE}
# Pre-compute PIT values and LOO intervals for these plots
# (avoids needing the loo and rstantools packages)
set.seed(456)
pit_vals <- runif(length(y))  # simulated PIT values for demonstration

# Pre-computed LOO predictive intervals (5 columns: ll, l, m, h, hh)
loo_intervals <- cbind(
  ll = apply(yrep, 2, quantile, probs = 0.05),
  l  = apply(yrep, 2, quantile, probs = 0.25),
  m  = apply(yrep, 2, quantile, probs = 0.50),
  h  = apply(yrep, 2, quantile, probs = 0.75),
  hh = apply(yrep, 2, quantile, probs = 0.95)
)
```

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_loo_pit_overlay, echo=FALSE, fig.show='hide'}
ppc_loo_pit_overlay(y, yrep, pit = pit_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_loo_pit_overlay", "PPC-loo.html", "ppc_loo_pit_overlay")
```

```{r ppc_loo_pit_qq, echo=FALSE, fig.show='hide'}
ppc_loo_pit_qq(y, yrep, pit = pit_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_loo_pit_qq", "PPC-loo.html", "ppc_loo_pit_qq")
```

```{r ppc_loo_intervals, echo=FALSE, fig.show='hide'}
ppc_loo_intervals(y, yrep, intervals = loo_intervals, order = "median")
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_loo_intervals", "PPC-loo.html", "ppc_loo_intervals")
```

```{r ppc_loo_ribbon, echo=FALSE, fig.show='hide'}
ppc_loo_ribbon(y, yrep, intervals = loo_intervals)
```

```{r results='asis', echo=FALSE}
gallery_card("ppc_loo_ribbon", "PPC-loo.html", "ppc_loo_ribbon")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Censoring {#ppc-censoring}

```{r setup-censoring, echo=FALSE}
# Create survival-type data for censoring plots
has_survival <- requireNamespace("survival", quietly = TRUE)
if (has_survival) {
  set.seed(789)
  n_cens <- 100
  y_cens <- rexp(n_cens, rate = 0.5)
  status_cens <- rbinom(n_cens, 1, prob = 0.7)
  y_cens[status_cens == 0] <- runif(sum(status_cens == 0), 0, y_cens[status_cens == 0])
  yrep_cens <- matrix(rexp(n_cens * 50, rate = 0.5), nrow = 50)
  group_cens <- factor(rep(c("A", "B"), each = n_cens / 2))
}
has_survival <- requireNamespace("survival", quietly = TRUE)
```

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppc_km_overlay, echo=FALSE, fig.show='hide', eval=has_survival}
ppc_km_overlay(y_cens, yrep_cens, status_y = status_cens)
```

```{r results='asis', echo=FALSE}
gallery_card_cond("ppc_km_overlay", "PPC-censoring.html", "ppc_km_overlay", has_survival)
```

```{r ppc_km_overlay_grouped, echo=FALSE, fig.show='hide', eval=has_survival}
ppc_km_overlay_grouped(y_cens, yrep_cens, status_y = status_cens, group = group_cens)
```

```{r results='asis', echo=FALSE}
gallery_card_cond("ppc_km_overlay_grouped", "PPC-censoring.html", "ppc_km_overlay_grouped", has_survival)
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```


# PPD: Posterior/Prior Predictive Distributions {#ppd}

Plots of posterior (or prior) predictive distributions *without*
comparing to observed data. These are the counterparts of the PPC
functions above but take only `ypred` rather than both `y` and `yrep`.
See `help("PPD-overview")` for details.

## Distributions {#ppd-distributions}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppd_hist, echo=FALSE, fig.show='hide'}
ppd_hist(yrep[1:5, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_hist", "PPD-distributions.html", "ppd_hist")
```

```{r ppd_dens_overlay, echo=FALSE, fig.show='hide'}
ppd_dens_overlay(yrep[1:25, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_dens_overlay", "PPD-distributions.html", "ppd_dens_overlay")
```

```{r ppd_ecdf_overlay, echo=FALSE, fig.show='hide'}
ppd_ecdf_overlay(yrep[1:25, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_ecdf_overlay", "PPD-distributions.html", "ppd_ecdf_overlay")
```

```{r ppd_dens, echo=FALSE, fig.show='hide'}
ppd_dens(yrep[200:202, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_dens", "PPD-distributions.html", "ppd_dens")
```

```{r ppd_freqpoly, echo=FALSE, fig.show='hide'}
ppd_freqpoly(yrep[1:3, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_freqpoly", "PPD-distributions.html", "ppd_freqpoly")
```

```{r ppd_boxplot, echo=FALSE, fig.show='hide'}
ppd_boxplot(yrep[1:8, ])
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_boxplot", "PPD-distributions.html", "ppd_boxplot")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Intervals {#ppd-intervals}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppd_intervals, echo=FALSE, fig.show='hide'}
ppd_intervals(yrep, x = x)
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_intervals", "PPD-intervals.html", "ppd_intervals")
```

```{r ppd_ribbon, echo=FALSE, fig.show='hide'}
ppd_ribbon(yrep, x = x)
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_ribbon", "PPD-intervals.html", "ppd_ribbon")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Test Statistics {#ppd-test-statistics}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r ppd_stat, echo=FALSE, fig.show='hide'}
ppd_stat(yrep, stat = "mean")
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_stat", "PPD-test-statistics.html", "ppd_stat")
```

```{r ppd_stat_grouped, echo=FALSE, fig.show='hide'}
ppd_stat_grouped(yrep, group = group, stat = "median")
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_stat_grouped", "PPD-test-statistics.html", "ppd_stat_grouped")
```

```{r ppd_stat_freqpoly, echo=FALSE, fig.show='hide'}
ppd_stat_freqpoly(yrep, stat = "mean")
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_stat_freqpoly", "PPD-test-statistics.html", "ppd_stat_freqpoly")
```

```{r ppd_stat_2d, echo=FALSE, fig.show='hide'}
ppd_stat_2d(yrep, stat = c("mean", "sd"))
```

```{r results='asis', echo=FALSE}
gallery_card("ppd_stat_2d", "PPD-test-statistics.html", "ppd_stat_2d")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```


# MCMC: Markov Chain Monte Carlo {#mcmc}

Plotting functions for MCMC draws and diagnostics.
See `help("MCMC-overview")` for details.

## Intervals & Areas {#mcmc-intervals}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_intervals, echo=FALSE, fig.show='hide'}
mcmc_intervals(draws)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_intervals", "MCMC-intervals.html", "mcmc_intervals")
```

```{r mcmc_areas, echo=FALSE, fig.show='hide'}
mcmc_areas(draws, prob = 0.8)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_areas", "MCMC-intervals.html", "mcmc_areas")
```

```{r mcmc_areas_ridges, echo=FALSE, fig.show='hide'}
mcmc_areas_ridges(draws)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_areas_ridges", "MCMC-intervals.html", "mcmc_areas_ridges")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Distributions {#mcmc-distributions}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_hist, echo=FALSE, fig.show='hide'}
mcmc_hist(draws)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_hist", "MCMC-distributions.html", "mcmc_hist")
```

```{r mcmc_hist_by_chain, echo=FALSE, fig.show='hide'}
mcmc_hist_by_chain(draws, pars = "alpha")
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_hist_by_chain", "MCMC-distributions.html", "mcmc_hist_by_chain")
```

```{r mcmc_dens, echo=FALSE, fig.show='hide'}
mcmc_dens(draws)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_dens", "MCMC-distributions.html", "mcmc_dens")
```

```{r mcmc_dens_overlay, echo=FALSE, fig.show='hide'}
mcmc_dens_overlay(draws, pars = c("alpha", "sigma"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_dens_overlay", "MCMC-distributions.html", "mcmc_dens_overlay")
```

```{r mcmc_dens_chains, echo=FALSE, fig.show='hide'}
mcmc_dens_chains(draws, pars = "alpha")
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_dens_chains", "MCMC-distributions.html", "mcmc_dens_chains")
```

```{r mcmc_violin, echo=FALSE, fig.show='hide'}
mcmc_violin(draws)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_violin", "MCMC-distributions.html", "mcmc_violin")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Traces & Rank Plots {#mcmc-traces}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_trace, echo=FALSE, fig.show='hide'}
color_scheme_set("mix-blue-red")
mcmc_trace(draws, pars = c("alpha", "sigma"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_trace", "MCMC-traces.html", "mcmc_trace")
```

```{r mcmc_trace_highlight, echo=FALSE, fig.show='hide'}
mcmc_trace_highlight(draws, pars = "alpha", highlight = 1)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_trace_highlight", "MCMC-traces.html", "mcmc_trace_highlight")
```

```{r mcmc_rank_overlay, echo=FALSE, fig.show='hide'}
color_scheme_set("brightblue")
mcmc_rank_overlay(draws, pars = "alpha")
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_rank_overlay", "MCMC-traces.html", "mcmc_rank_overlay")
```

```{r mcmc_rank_hist, echo=FALSE, fig.show='hide'}
mcmc_rank_hist(draws, pars = "alpha")
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_rank_hist", "MCMC-traces.html", "mcmc_rank_hist")
```

```{r mcmc_rank_ecdf, echo=FALSE, fig.show='hide'}
mcmc_rank_ecdf(draws, pars = "alpha")
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_rank_ecdf", "MCMC-traces.html", "mcmc_rank_ecdf")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Scatterplots {#mcmc-scatterplots}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_scatter, echo=FALSE, fig.show='hide'}
color_scheme_set("brightblue")
mcmc_scatter(draws, pars = c("alpha", "sigma"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_scatter", "MCMC-scatterplots.html", "mcmc_scatter")
```

```{r mcmc_hex, echo=FALSE, fig.show='hide'}
mcmc_hex(draws, pars = c("alpha", "sigma"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_hex", "MCMC-scatterplots.html", "mcmc_hex")
```

```{r mcmc_pairs, echo=FALSE, fig.show='hide'}
mcmc_pairs(draws, pars = c("alpha", "sigma", "beta[1]"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_pairs", "MCMC-scatterplots.html", "mcmc_pairs")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Diagnostics {#mcmc-diagnostics}

```{r setup-diagnostics, echo=FALSE}
# Create example diagnostic values
set.seed(321)
rhat_vals <- c(1.00, 1.01, 1.02, 1.005, 1.08, 1.001)
names(rhat_vals) <- dimnames(draws)[[3]]
neff_vals <- c(0.9, 0.8, 0.7, 0.85, 0.5, 0.95)
names(neff_vals) <- dimnames(draws)[[3]]
```

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_rhat, echo=FALSE, fig.show='hide'}
mcmc_rhat(rhat_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_rhat", "MCMC-diagnostics.html", "mcmc_rhat")
```

```{r mcmc_rhat_hist, echo=FALSE, fig.show='hide'}
mcmc_rhat_hist(rhat_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_rhat_hist", "MCMC-diagnostics.html", "mcmc_rhat_hist")
```

```{r mcmc_neff, echo=FALSE, fig.show='hide'}
mcmc_neff(neff_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_neff", "MCMC-diagnostics.html", "mcmc_neff")
```

```{r mcmc_neff_hist, echo=FALSE, fig.show='hide'}
mcmc_neff_hist(neff_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_neff_hist", "MCMC-diagnostics.html", "mcmc_neff_hist")
```

```{r mcmc_acf, echo=FALSE, fig.show='hide'}
mcmc_acf(draws, pars = c("alpha", "sigma"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_acf", "MCMC-diagnostics.html", "mcmc_acf")
```

```{r mcmc_acf_bar, echo=FALSE, fig.show='hide'}
mcmc_acf_bar(draws, pars = c("alpha", "sigma"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_acf_bar", "MCMC-diagnostics.html", "mcmc_acf_bar")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Parallel Coordinates {#mcmc-parcoord}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_parcoord, echo=FALSE, fig.show='hide'}
mcmc_parcoord(draws, pars = c("alpha", "sigma", "beta[1]", "beta[2]"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_parcoord", "MCMC-parcoord.html", "mcmc_parcoord")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Recovery {#mcmc-recover}

```{r setup-recover, echo=FALSE}
# Create example "true" parameter values for recovery plots
true_vals <- c(alpha = 2.5, sigma = 1.2, "beta[1]" = 0.5, "beta[2]" = -0.3,
               "beta[3]" = 0.8, "beta[4]" = -1.1)
```

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_recover_intervals, echo=FALSE, fig.show='hide'}
mcmc_recover_intervals(draws, true = true_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_recover_intervals", "MCMC-recover.html", "mcmc_recover_intervals")
```

```{r mcmc_recover_scatter, echo=FALSE, fig.show='hide'}
mcmc_recover_scatter(draws, true = true_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_recover_scatter", "MCMC-recover.html", "mcmc_recover_scatter")
```

```{r mcmc_recover_hist, echo=FALSE, fig.show='hide'}
mcmc_recover_hist(draws, true = true_vals)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_recover_hist", "MCMC-recover.html", "mcmc_recover_hist")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```

## Combo Plots {#mcmc-combo}

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_combo, echo=FALSE, fig.show='hide', fig.width=4.5, fig.height=4}
mcmc_combo(draws, pars = c("alpha", "sigma"))
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_combo", "MCMC-combos.html", "mcmc_combo")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```


# HMC/NUTS Diagnostics {#nuts}

Diagnostic plots specific to Hamiltonian Monte Carlo (HMC) and the No-U-Turn
Sampler (NUTS). These functions require NUTS-specific diagnostic information.
See `help("MCMC-nuts")` for details.

```{r setup-nuts, echo=FALSE}
# Simulate simple NUTS diagnostic data
set.seed(654)
n_iter <- 250
n_chains <- 4
np_data <- data.frame(
  Parameter = rep(c("accept_stat__", "stepsize__", "treedepth__",
                     "n_leapfrog__", "divergent__", "energy__"), each = n_iter * n_chains),
  Value = c(
    rbeta(n_iter * n_chains, 20, 1),         # accept_stat__
    rep(c(0.8, 0.9, 0.85, 0.75), each = n_iter),  # stepsize__
    sample(2:6, n_iter * n_chains, replace = TRUE, prob = c(0.05, 0.3, 0.4, 0.2, 0.05)),  # treedepth__
    2^sample(2:6, n_iter * n_chains, replace = TRUE, prob = c(0.05, 0.3, 0.4, 0.2, 0.05)),  # n_leapfrog__
    rbinom(n_iter * n_chains, 1, 0.02),      # divergent__
    rnorm(n_iter * n_chains, -200, 5)        # energy__
  ),
  Chain = rep(rep(1:n_chains, each = n_iter), 6),
  Iteration = rep(rep(1:n_iter, n_chains), 6)
)

lp_data <- data.frame(
  Value = rnorm(n_iter * n_chains, -200, 2),
  Chain = rep(1:n_chains, each = n_iter),
  Iteration = rep(1:n_iter, n_chains)
)
```

```{r results='asis', echo=FALSE}
cat('<div class="gallery-grid">\n')
```

```{r mcmc_nuts_divergence, echo=FALSE, fig.show='hide'}
mcmc_nuts_divergence(np_data, lp_data)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_nuts_divergence", "MCMC-nuts.html", "mcmc_nuts_divergence")
```

```{r mcmc_nuts_energy, echo=FALSE, fig.show='hide'}
mcmc_nuts_energy(np_data)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_nuts_energy", "MCMC-nuts.html", "mcmc_nuts_energy")
```

```{r mcmc_nuts_acceptance, echo=FALSE, fig.show='hide'}
mcmc_nuts_acceptance(np_data, lp_data)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_nuts_acceptance", "MCMC-nuts.html", "mcmc_nuts_acceptance")
```

```{r mcmc_nuts_treedepth, echo=FALSE, fig.show='hide'}
mcmc_nuts_treedepth(np_data, lp_data)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_nuts_treedepth", "MCMC-nuts.html", "mcmc_nuts_treedepth")
```

```{r mcmc_nuts_stepsize, echo=FALSE, fig.show='hide'}
mcmc_nuts_stepsize(np_data, lp_data)
```

```{r results='asis', echo=FALSE}
gallery_card("mcmc_nuts_stepsize", "MCMC-nuts.html", "mcmc_nuts_stepsize")
```

```{r results='asis', echo=FALSE}
cat('</div>\n')
```
