---
title: "Plotting MCMC draws using the bayesplot package"
author: "Jonah Gabry"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Plotting MCMC draws}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, settings, include=FALSE}
library(bayesplot)
knitr::opts_chunk$set(
  dev = "pdf",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  comment = NA
)
```

This vignette focuses on plotting parameter estimates from MCMC draws. MCMC
diagnostic plots are covered in the separate vignette [*Visual MCMC diagnostics
using the bayesplot package*]("MCMC-diagnostics.html").

## Plots for MCMC draws

The **bayesplot** package provides various plotting functions for visualizing 
Markov chain Monte Carlo (MCMC) draws from the posterior distribution of the 
parameters of a Bayesian model.

In this vignette we'll use draws obtained using the `stan_glm` function in the 
**rstanarm** package (Gabry and Goodrich, 2016), but MCMC draws from using 
any package can be used with the functions in the **bayesplot** package. See,
for example, **brms** (which, like **rstanarm**, calls the **rstan** package
internally to use [Stan](http://mc-stan.org/)'s MCMC sampler).

```{r, eval=FALSE, results='hide'}
library("rstanarm")
fit <- stan_glm(
  mpg ~ ., 
  data = mtcars, 
  chains = 4, 
  iter = 2000,
  seed = 1111
)
print(fit)
```

```{r stan_glm, echo=FALSE, results='hide'}
suppressPackageStartupMessages(library("rstanarm"))
fit <- stan_glm(
  mpg ~ ., data = mtcars, 
  chains = 4, 
  iter = 2000,
  seed = 1111
)
```

```{r, print-fit, echo=FALSE}
print(fit)
```

To use the posterior draws with the functions in the **bayesplot** package
we'll extract them from the fitted model object: 

```{r, get-draws}
posterior <- as.array(fit)
dim(posterior)
dimnames(posterior)
```

### Interval estimates

Posterior intervals for the parameters can be plotted using the `mcmc_intervals`
function.

```{r, mcmc_intervals}
library("bayesplot")
mcmc_intervals(posterior, pars = c("cyl", "drat", "am", "sigma"))
```

The default is to show 50% intervals (the thick lines) and 90% intervals 
(the thinner outer lines above). These defaults can be changed using the `prob`
and `prob_outer` arguments, respectively. The points in the above plot 
are posterior medians. The `point_est` argument can be used to select posterior
means instead or to omit the point estimates.

To show the uncertainty intervals as shaded areas under the estimated posterior
density curves we can use the `mcmc_areas` function:

```{r, mcmc_areas}
mcmc_areas(
  posterior, 
  pars = c("cyl", "drat", "am", "sigma"),
  prob = 0.8, # 80% intervals
  prob_outer = 0.99, # 99%
  point_est = "mean"
)
```


### Histograms and density estimates

Histograms or kernel density estimates of posterior distributions of the various
model parameters can be visualized using the functions described on the 
`MCMC-distributions` page in the **bayesplot** documentation.

#### Histograms

First we'll use the `mcmc_hist` function to plot the posterior distributions
of several of the regression coefficients:

```{r, mcmc_hist, message=FALSE}
mcmc_hist(posterior, pars = c("wt", "am"))
```

In the plot above the four Markov chains are merged. To view the chains 
separately we can use the `mcmc_hist_by_chain` function:

```{r, mcmc_hist_by_chain, message=FALSE}
mcmc_hist_by_chain(posterior, pars = c("wt", "am"))
```

#### Densities

The `mcmc_dens` function is the same as `mcmc_hist` except it plots 
kernel density estimates rather than histograms:

```{r, mcmc_dens}
mcmc_dens(posterior, pars = c("wt", "am"))
```

We can also view density estimates for the individual chains overlaid on the 
same plot using the `mcmc_dens_overlay` function:

```{r, mcmc_dens_overlay}
mcmc_dens_overlay(posterior, pars = c("wt", "am"))
```

And the `mcmc_violin` function plots the density estimates of each chain as
violins with horizontal lines at notable quantiles:

```{r, mcmc_violin}
mcmc_violin(posterior, pars = c("wt", "am"), probs = c(0.1, 0.5, 0.9))
```

### Scatterplots

The `mcmc_scatter` function creates a scatterplot with two parameters:

```{r, mcmc_scatter}
mcmc_scatter(posterior, pars = c("(Intercept)", "wt"), size = 1.5, alpha = 0.5)
```

In addition to `mcmc_scatter`, an `mcmc_pairs` functions for creating pairs
plots with more than two parameters will be available in one of the next
releases of the package.

### Traceplots 

The `mcmc_trace` function creates standard traceplots:

```{r, mcmc_trace}
mcmc_trace(posterior, pars = c("wt", "sigma"))
```

The `mcmc_trace_highlight` uses points instead of lines and reduces the opacity
of all but a single chain.

```{r, mcmc_trace_highlight}
mcmc_trace_highlight(
  posterior, 
  pars = c("wt", "sigma"), 
  highlight = 3 # highlight the third chain by reducing opacity of others
)
```

## References

Gabry, J., and Goodrich, B. (2016). rstanarm: Bayesian Applied Regression 
Modeling via Stan. R package version 2.11.1. 
http://mc-stan.org/interfaces/rstanarm.html

Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., and
Rubin, D. B. (2013). *Bayesian Data Analysis*. Chapman & Hall/CRC Press, London,
third edition.

Stan Development Team. (2016). *Stan Modeling Language Users
Guide and Reference Manual*. http://mc-stan.org/documentation/
