---
title: "Visual MCMC diagnostics using the bayesplot package"
author: "Jonah Gabry"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Visual MCMC diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, settings, include=FALSE}
library("bayesplot")
knitr::opts_chunk$set(
  dev = "svg",
  fig.align = "center",
  fig.width = 6,
  fig.height = 4
)
```

## Overview

A Markov chain generates samples from the target distribution only after it has
converged to equilibrium. Unfortunately, this is only guaranteed in the limit
in theory. In practice, diagnostics must be applied to monitor whether the
Markov chain(s) have converged.

The __bayesplot__ package provides various plotting functions for visualizing 
Markov chain Monte Carlo (MCMC) diagnostics after fitting a Bayesian model.

## General MCMC diagnostics

#### Rhat: potential scale reduction statistic

> One way to monitor whether a chain has converged to the equilibrium distribution
is to compare its behavior to other randomly initialized chains. This is the 
motivation for the Gelman and Rubin (1992) potential scale reduction statistic, 
$\hat{R}$. The $\hat{R}$ statistic measures the ratio of the average variance of
samples within each chain to the variance of the pooled samples across chains;
if all chains are at equilibrium, these will be the same and $\hat{R}$ will be
one. If the chains have not converged to a common distribution, the $\hat{R}$
statistic will be greater than one. ([Stan Modeling Language Users Guide and Reference Manual](http://mc-stan.org/documentation/))

The **bayesplot** package provides the functions `mcmc_rhat` and
`mcmc_rhat_hist` for visualizing $\hat{R}$ estimates. To demostrate, 
we'll use the `stan_glm` function in the **rstanarm** package to quickly 
get some real $\hat{R}$ values to work with.

First we'll intentionally use too few MCMC iterations, which should 
lead to some high $\hat{R}$ values (we'll probably get convergence
warnings from **rstanarm**).

```{r, eval=FALSE, results='hide'}
library("rstanarm")
fit <- stan_glm(
  mpg ~ ., data = mtcars, 
  chains = 4, iter = 50, # intentionally too few iterations
  seed = 1111
)
```

```{r stan_glm, echo=FALSE, results='hide'}
suppressPackageStartupMessages(library("rstanarm"))
fit <- stan_glm(
  mpg ~ ., data = mtcars, 
  chains = 4, 
  iter = 50,
  seed = 1111
)
```

The **bayesplot** package provides a generic `rhat` extractor function, 
currently with methods defined for models fit using the **rstan** and
**rstanarm** packages. But regardless of how you fit your model, all 
**bayesplot** needs is a vector of $\hat{R}$ values.

```{r print-rhats}
library("bayesplot")
rhats <- rhat(fit)
print(rhats)
```

We can visualize the $\hat{R}$ values with the `mcmc_rhat` function:

```{r mcmc_rhat}
mcmc_rhat(rhats)
```

In the plot, the points representing the $\hat{R}$ values are colored based on
whether they are less than $1.05$, between $1.05$ and $1.1$, or greater than
$1.1$.

We can see which parameters have the concerning $\hat{R}$ values by turning
on the $y$-axis text using the `yaxis_text` convenience function:

```{r, mcmc_rhat-2}
mcmc_rhat(rhats) + yaxis_text()
```

The axis $y$-axis text if off by default for this plot because it's only 
possible to see the labels for models like this one with very few parameters.

If we refit the model using longer Markov chains we should see all $\hat{R} <
1.1$, and all points in the plot the same (light) color:

```{r, results='hide'}
fit2 <- stan_glm(
  mpg ~ ., data = mtcars, 
  chains = 4, iter = 1000,
  seed = 1111
)

mcmc_rhat(rhat = rhat(fit2))
```



We can see the same information shown by `mcmc_rhat` but in histogram form using
the `mcmc_rhat_hist` function. See the **Examples** section in
`help("mcmc_rhat_hist")` for examples.


#### Effective sample size

Write this section.

## Algorithm specific MCMC diagnostics

#### Diagnostics for the No-U-Turn sampler (NUTS)

Write this section.



## References

Gabry, J., and Goodrich, B. (2016). rstanarm: Bayesian Applied Regression 
Modeling via Stan. R package version `r packageVersion("rstanarm")`.
http://mc-stan.org/interfaces/rstanarm.html

Gelman, A. and Rubin, D. B. (1992). Inference from iterative simulation using
multiple sequences. *Statistical Science*. 7(4): 457--472.

Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., and
Rubin, D. B. (2013). *Bayesian Data Analysis*. Chapman & Hall/CRC Press, London,
third edition.

Stan Development Team. (2016). *Stan Modeling Language Users
Guide and Reference Manual*. http://mc-stan.org/documentation/

