---
title: "Visual MCMC diagnostics using the bayesplot package"
author: "Jonah Gabry"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Visual MCMC diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, settings, include=FALSE}
library("bayesplot")
library("ggplot2")
library("gridExtra")
library("rstan")
knitr::opts_chunk$set(
  dev = "svg",
  fig.align = "center",
  fig.width = 4,
  fig.height = 3,
  comment = NA
)
```

This vignette focuses on MCMC diagnostic plots. Plots of parameter estimates
from MCMC draws are covered in the separate vignette
[*Plotting MCMC draws using the bayesplot package*]("MCMC.html"), 
and graphical posterior predictive checks are covered in 
[*Graphical posterior predictive checks using the bayesplot package*]("PPC.rmd").

In addition to __bayesplot__ we'll load the following packages: 

* __ggplot2__ for customizing the ggplot objects created by __bayesplot__
* __gridExtra__ for displaying multiple ggplot objects in a grid
* __rstan__ for fitting the example models used throughout the vignette

```{r, eval=FALSE}
library("bayesplot")
library("ggplot2")    
library("gridExtra") 
library("rstan")      
```

## General MCMC diagnostics

A Markov chain generates draws from the target distribution only after it has 
converged to equilibrium. Unfortunately, this is only guaranteed in the limit in
theory. In practice, diagnostics must be applied to monitor whether the Markov
chain(s) have converged. The __bayesplot__ package provides various plotting
functions for visualizing Markov chain Monte Carlo (MCMC) diagnostics after
fitting a Bayesian model. MCMC draws from any package can be used with
__bayesplot__, although there are a few diagnostic plots that are specifically
intended to be used for [Stan](http://mc-stan.org/) models.

To demonstrate, in this vignette we'll use the eight schools example discussed
in  Rubin (1981), Gelman et al (2013), and the 
[RStan Getting Started](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started#how-to-use-rstan)
wiki. This is a simple hierarchical meta-analysis model with data consisting of 
point estimates `y` and standard errors `sigma` from analyses of test prep 
programs in `J=8` schools:

```{r, schools_dat}
schools_dat <- list(
  J = 8, 
  y = c(28,  8, -3,  7, -1,  1, 18, 12),
  sigma = c(15, 10, 16, 11,  9, 11, 10, 18)
)
```

The model is: 
$$
\begin{align*}
y_j &\sim {\rm Normal}(\theta_j, \sigma_j), \quad j = 1,\dots,J \\
\theta_j &\sim {\rm Normal}(\mu, \tau), \quad j = 1, \dots, J \\
\mu &\sim {\rm Normal}(0, 10) \\
\tau &\sim {\rm half-Cauchy}(0, 10),
\end{align*}
$$
with the normal distribution parameterized by the mean and standard deviation, 
not the variance or precision. In Stan code:
```{stan, stancode1, output.var = "schools_mod1"}
// Saved in 'schools_mod1.stan'
data {
  int<lower=0> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}
parameters {
  real mu;
  real<lower=0> tau;
  vector[J] theta;
}
model {
  mu ~ normal(0, 10);
  tau ~ cauchy(0, 10);
  theta ~ normal(mu, tau);
  y ~ normal(theta, sigma);
}
```

This parameterization of the model is referred to as the centered 
parameterization (CP). We'll also fit the same statistical model but using the 
so-called non-centered parameterization (NCP), which replaces the vector 
$\theta$ with a vector $\eta$ of a priori _i.i.d._ standard normal parameters 
and then contructs $\theta$ deterministically from $\eta$ by scaling by $\tau$ 
and shifting by $\mu$:
$$
\begin{align*}
\theta_j &= \mu + \tau \,\eta_j, \quad j = 1,\dots,J \\
\eta_j &\sim N(0,1), \quad j = 1,\dots,J.
\end{align*}
$$
The Stan code for this model is:
```{stan, stancode2, output.var = "schools_mod2"}
// Saved in 'schools_mod2.stan'
data {
  int<lower=0> J;
  vector[J] y;
  vector<lower=0>[J] sigma;
}
parameters {
  real mu;
  real<lower=0> tau;
  vector[J] eta;
}
transformed parameters {
  vector[J] theta;
  theta = mu + tau * eta;
}
model {
  mu ~ normal(0, 10);
  tau ~ cauchy(0, 10);
  eta ~ normal(0,1); // implies theta ~ normal(mu, tau)
  y ~ normal(theta, sigma);
}
```
The centered and non-centered are two parameterizations of the same statistical 
model, but they have very different practical implications for MCMC. Using the 
__bayesplot__ diagnostic plots, we'll see that, for this data, the NCP is 
required in order to properly explore the posterior distribution.

To fit both models we first translate the Stan code to C++ and compile it using 
the `stan_model` function.
```{r, compile-models, eval=FALSE}
schools_mod1 <- stan_model("schools_mod1.stan")
schools_mod2 <- stan_model("schools_mod2.stan")
```
We then fit the model by calling Stan's MCMC algorithm using the `sampling` 
function,
```{r, fit-models-hidden, results='hide', message=FALSE}
fit1 <- sampling(schools_mod1, data = schools_dat)
fit2 <- sampling(schools_mod2, data = schools_dat)
```
and extract a `iterations x chains x parameters` array of posterior draws with 
`as.array`,
```{r, extract-draws}
# Extract posterior draws for later use
posterior1 <- as.array(fit1)
posterior2 <- as.array(fit2)
```

For now ignore any warnings about divergent transitions after warmup. We will
come back to those later in the vignette in the [Diagnostics for the No-U-Turn
Sampler](#diagnostics-for-the no-u-turn-sampler) section.


### Rhat: potential scale reduction statistic

> One way to monitor whether a chain has converged to the equilibrium distribution
is to compare its behavior to other randomly initialized chains. This is the
motivation for the Gelman and Rubin (1992) potential scale reduction statistic,
$\hat{R}$. The $\hat{R}$ statistic measures the ratio of the average variance of
samples within each chain to the variance of the pooled samples across chains;
if all chains are at equilibrium, these will be the same and $\hat{R}$ will be
one. If the chains have not converged to a common distribution, the $\hat{R}$
statistic will be greater than one. 
(Stan Development Team, 2016).

The **bayesplot** package provides the functions `mcmc_rhat` and
`mcmc_rhat_hist` for visualizing $\hat{R}$ estimates.

First we'll quickly fit one of the models above again, this time intentionally
using too few MCMC iterations. This should lead to some high $\hat{R}$ values.

```{r, results='hide'}
fit1_50iter <- sampling(schools_mod1, data = schools_dat, chains = 2, iter = 50)
```

 **bayesplot** also provides a generic `rhat` extractor function,
currently with methods defined for models fit using the **rstan** and
**rstanarm** packages. But regardless of how you fit your model, all
**bayesplot** needs is a vector of $\hat{R}$ values.

```{r print-rhats}
library("bayesplot")
rhats <- rhat(fit1_50iter)
print(rhats)
```

We can visualize the $\hat{R}$ values with the `mcmc_rhat` function:

```{r mcmc_rhat}
color_scheme_set("brightblue") # see help("color_scheme_set")
mcmc_rhat(rhats)
```

In the plot, the points representing the $\hat{R}$ values are colored based on
whether they are less than $1.05$, between $1.05$ and $1.1$, or greater than
$1.1$.

We can see the names of the parameters with the concerning $\hat{R}$ values by
turning on the $y$-axis text using the `yaxis_text` convenience function:

```{r, mcmc_rhat-2}
mcmc_rhat(rhats) + yaxis_text()
```

The axis $y$-axis text is off by default for this plot because it's only
possible to see the labels clearly for models with very few parameters.

If we look at the same model fit using longer Markov chains we should see all
$\hat{R} < 1.1$, and all points in the plot the same (light) color:

```{r, results='hide'}
mcmc_rhat(rhat = rhat(fit1)) + yaxis_text()
```

We can see the same information shown by `mcmc_rhat` but in histogram form using
the `mcmc_rhat_hist` function. See the **Examples** section in 
`help("mcmc_rhat_hist")` for examples.


### Effective sample size

The effective sample size is an estimate of the number of independent draws from
the posterior distribution of the estimand of interest. Because the draws within
a Markov chain are _not_ independent if there is autocorrelation, the effective 
sample size, $n_{eff}$, will be smaller than the total sample size, $N$. The 
larger the ratio of $n_{eff}$ to $N$ the better.

The **bayesplot** package provides a generic `neff_ratio` extractor function,
currently with methods defined for models fit using the **rstan** and
**rstanarm** packages. But regardless of how you fit your model, all
**bayesplot** needs is a vector of $n_{eff}/N$ values. The `mcmc_neff` and `mcmc_neff_hist` can then be used to plot the ratios.

```{r print-neff-ratios}
ratios1 <- neff_ratio(fit1)
print(ratios1)
mcmc_neff(ratios1)
```

In the plot, the points representing the values of $n_{eff}/N$ are colored based
on whether they are less than $0.1$, between $0.1$ and $0.5$, or greater than
$0.5$. These particular values are arbitrary in that they have no particular
theoretical meaning, but a useful heuristic is to worry about any $n_{eff}/N$
less than $0.1$.

One important thing to keep in mind is that these ratios will depend not only on
the model being fit but also on the particular MCMC algorithm used. One reason
why we have such high ratios of $n_{eff}$ to $N$ is that the No-U-Turn sampler
used by **rstan** generally produces draws from the posterior distribution with
much lower autocorrelations compared to draws obtained using other MCMC
algorithms (e.g., Gibbs).

Even for models fit using **rstan** the parameterization can make a big
difference. Here are the $n_{eff}/N$ plots for `fit1` and `fit2` side by side.

```{r mcmc_neff-compare}
# A function we'll use several times to plot comparisons of the centered
# parameterization (cp) and the non-centered parameterization (ncp)
compare_ <- function(cp_plot, ncp_plot) {
  grid.arrange(
    cp_plot + labs(subtitle = "Centered parameterization"),
    ncp_plot + labs(subtitle = "Non-centered parameterization")
  )
}

neff1 <- neff_ratio(fit1, pars = c("theta", "mu", "tau"))
neff2 <- neff_ratio(fit2, pars = c("theta", "mu", "tau"))
compare_parameterizations(mcmc_neff(neff1), mcmc_neff(neff2))
```

Because of the difference in parameterization, the effective sample sizes are 
much better for the second model, the non-centered parameterization.

### Autocorrelation

As mentioned above, $n_{eff}/N$ decreases as autocorrelation becomes more 
extreme. We can visualize the autocorrelation using the `mcmc_acf` (line plot) 
or `mcmc_acf_bar` (bar plot) functions. For the selected parameters, these 
functions show the autocorrelation for each Markov chain separately up to a 
user-specified number of lags.

Here we can again see a difference when comparing the two parameterizations of 
the same model. For model 1, $\theta_1$ is the primitive parameter for school 1,
whereas for the non-centered parameterization in model 2 the primitive parameter
is $\eta_1$ (and $\theta_1$ is later constructed from $\eta_1$, $\mu$, and 
$\tau$):

```{r mcmc_acf}
grid.arrange(
  mcmc_acf(posterior1, pars = "theta[1]", lags = 10),
  mcmc_acf(posterior2, pars = "eta[1]", lags = 10),
  ncol = 2
)
```


## Diagnostics for the No-U-Turn Sampler

The No-U-Turn Sampler (NUTS, Hoffman and Gelman, 2014) is the variant of
Hamiltonian Monte Carlo (HMC) used by [Stan](http://mc-stan.org/) and the various R
packages that depend on Stan for fitting Bayesian models.

The **bayesplot** package has special functions for visualizing some of the
unique diagnostics permitted by HMC, and NUTS in particular. See Betancourt and
Girolami (2013), Betancourt (2016), and Stan Development Team (2016) for more
details on the concepts.

The special **bayesplot** functions for NUTS diagnostics are

* `mcmc_nuts_acceptance`
* `mcmc_nuts_divergence`
* `mcmc_nuts_stepsize`
* `mcmc_nuts_treedepth`
* `mcmc_nuts_energy`

The **bayesplot** package also provides generic functions `log_posterior` and 
`nuts_params` for extracting the required information for the plots from fitted
model objects. Currently methods are provided for models fit using the **rstan**
and **rstanarm** packages, although it is not difficult to define additional
methods for the objects returned by other R packages.

For the Stan models we fit above we can use the `log_posterior` and
`nuts_params` methods for stanfit objects:

```{r, extract-nuts-info}
lp1 <- log_posterior(fit1)
head(lp1)
np1 <- nuts_params(fit1)
head(np1)

# for the second model
lp2 <- log_posterior(fit2)
np2 <- nuts_params(fit2)
```

### Divergent transitions

When running the Stan models at the beginning of this vignette
there were warnings about divergent transitions.
For an explanation of these warnings see [Divergent transitions after warmup](http://mc-stan.org/misc/warnings.html#divergent-transitions-after-warmup).

To look at some information about the divergences we can use the
`mcmc_nuts_divergence` function:

```{r, mcmc_nuts_divergence}
color_scheme_set("red")
mcmc_nuts_divergence(np1, lp1)
```

In the top panel we see the distribution of the log-posterior when there was no
divergence vs the distribution when there was a divergence. Divergences often
indicate that some part of the posterior isn't being explored and we will see
that `lp|Divergence` has lighter tails than `lp|No divergence`.

The bottom panel shows the same thing but instead of the log-posterior the NUTS 
acceptance statistic is shown.

Specifying the optional `chain` argument will overlay the plot just for a
particular Markov chain on the plot for all chains combined:

```{r, mcmc_nuts_divergence-chain}
mcmc_nuts_divergence(np1, lp1, chain = 4)
```

For the non-centered parameterization we also had warnings about divergences

```{r, mcmc_nuts_divergence-2}
mcmc_nuts_divergence(np2, lp2)
```

but we have far fewer of them to worry about. If there are only a few 
divergences we can often get rid of them by increasing the target acceptance 
rate (`adapt_delta`), which has the effect of lowering the stepsize used by the 
sampler and allowing the Markov chains to explore more complicated curvature in
the target distribution.

```{r, fit-adapt-delta, results='hide',message=FALSE}
fit1b <- sampling(schools_mod1, data = schools_dat,
                  control = list(adapt_delta = 0.99))
fit2b <- sampling(schools_mod2, data = schools_dat,
                  control = list(adapt_delta = 0.99))
```

For the first model and this particular data, increasing
`adapt_delta` will not solve the problem and a reparameterization 
is required. 


```{r, mcmc_nuts_divergence-3}
mcmc_nuts_divergence(nuts_params(fit1b), log_posterior(fit1b))
mcmc_nuts_divergence(nuts_params(fit2b), log_posterior(fit2b))
```

### Energy and Bayesian fraction of missing information

The `mcmc_nuts_energy` function creates plots similar to those presented in
Betancourt (2016). While `mcmcm_nuts_divergence` can identify light tails and
incomplete exploration of the target distribution, the `mcmc_nuts_energy`
function can identify overly heavy tails that are also challenging for sampling.
Informally, the energy diagnostic for HMC (and the related energy-based Bayesian
fraction of missing information) quantify the heaviness of the tails of the
posterior distribution.

The plot created by `mcmc_nuts_energy` shows overlaid histograms of the 
(centered) marginal energy distribution $\pi_E$ and the first-differenced 
distribution $\pi_{\Delta E}$,
```{r, mcmc_nuts_energy-1, message=FALSE}
mcmc_nuts_energy(np1)
```
To keep the chains separate add `merge_chains=FALSE`:
```{r, mcmc_nuts_energy-2}
mcmc_nuts_energy(np1, merge_chains = FALSE, binwidth = 1)
```

The two histograms ideally look the same (Betancourt, 2016), which is only the
case for the non-centered parameterization (bottom panel):

```{r, mcmc_nuts_energy-3, message=FALSE}
grid.arrange(
  mcmc_nuts_energy(np1, binwidth = 1 / 2) + 
    labs(subtitle = "Centered parameterization"),
  mcmc_nuts_energy(np2, binwidth = 1 / 2) + 
    labs(subtitle = "Non-centered parameterization")
  )
```

The difference between the parameterizations is even more apparent if we force 
the stepsize to a smaller value and help the chains explore more of the 
posterior

```{r, mcmc_nuts_energy-4, message=FALSE}
np1b <- nuts_params(fit1b)
np2b <- nuts_params(fit2b)
grid.arrange(mcmc_nuts_energy(np1b), 
             mcmc_nuts_energy(np2b))
```

See Betancourt (2016) for more on this particular example as well as the general
theory behind the energy plots.

## References

In future versions of the **bayesplot** package this vignette will be expanded
to provide examples of interpreting all of the NUTS diagnostic plots. For
now see `help("MCMC-nuts", package = "bayesplot")` and the references below.

-----

Betancourt, M. (2016). Diagnosing suboptimal cotangent disintegrations in
Hamiltonian Monte Carlo. https://arxiv.org/abs/1604.00695

Betancourt, M. and Girolami, M. (2013). Hamiltonian Monte Carlo for hierarchical
models. https://arxiv.org/abs/1312.0906

Gabry, J., and Goodrich, B. (2016). rstanarm: Bayesian Applied Regression
Modeling via Stan. R package version 2.11.1.
http://mc-stan.org/interfaces/rstanarm.html

Gelman, A. and Rubin, D. B. (1992). Inference from iterative simulation using
multiple sequences. *Statistical Science*. 7(4): 457--472.

Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., and Rubin,
D. B. (2013). *Bayesian Data Analysis*. Chapman & Hall/CRC Press, London, third
edition.

Hoffman, M. D. and Gelman, A. (2014). The No-U-Turn Sampler: adaptively setting path lengths in Hamiltonian Monte Carlo. *Journal of Machine Learning Research*. 15:1593--1623.

Rubin, D. B. (1981). Estimation in Parallel Randomized Experiments. *Journal of
Educational and Behavioral Statistics*. 6:377–401.

Stan Development Team. (2016). *Stan Modeling Language Users
Guide and Reference Manual*. http://mc-stan.org/documentation/

