---
title: "PPC Calibration"
author: "bayesplot team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{PPC Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/SETTINGS-knitr.txt"}
```

```{r pkgs}
library("ggplot2")
library("patchwork")
library("rstanarm")
library("brms")

SEED <- 840
set.seed(SEED)
```

## Overview

This vignette introduces the **PPC calibration** family of functions in
`bayesplot`. Calibration plots let you assess whether a model's predicted
probabilities agree with observed event rates — a key aspect of predictive
model checking for binary outcomes.

A well-calibrated model produces predicted probabilities that match empirical
event frequencies. For example, among observations where the model predicts a
30 % chance of an event, roughly 30 % of those events should actually occur.
Deviations from this pattern reveal systematic over- or underprediction.

The sections below first explain the data structure returned by
`ppc_calibration_data()`, then walk through the plotting functions step by step.

---

## Simulated example data

Throughout this vignette we use a simple simulated binary outcome. We generate
400 observations from a logistic model, and construct two prediction systems:
one approximately calibrated and one deliberately miscalibrated.

```{r simulate}
# Simulate binary outcome
N <- 400
S <- 80
x <- runif(N, -2, 0)

p_true <- plogis(0.8 * x)
y      <- rbinom(N, size = 1, prob = p_true)

# Helper: generate S posterior-like draws of predicted probabilities.
# Noise is added on the logit scale so predictions remain in (0, 1).
make_prep <- function(mu, S = 100, sd_logit = 0.35) {
  lp  <- qlogis(pmin(pmax(mu, 1e-6), 1 - 1e-6))
  mat <- matrix(
    rnorm(S * length(mu), mean = rep(lp, each = S), sd = sd_logit),
    nrow = S, byrow = FALSE
  )
  plogis(mat)
}

# Two prediction systems
# Nearly calibrated: centred at the true probabilities
prep_good <- make_prep(p_true, S = S, sd_logit = 0.30)

# Miscalibrated: probabilities pushed to the extremes and shifted
p_bad_center <- plogis(1.8 * qlogis(p_true) - 0.5)
prep_bad     <- make_prep(p_bad_center, S = S, sd_logit = 0.25)
```

---

## `ppc_calibration_data()` - the underlying data structure

`ppc_calibration_data()` computes the data that all calibration plots are built
on. Understanding its output helps you work with or extend the visualisations.

```{r calib-data}
dat <- ppc_calibration_data(y, prep_good, group = NULL)
head(dat, 5)
```

The returned data frame has the following columns:

| Column | Description |
|---|---|
| `group` | Group label $z = 1, \ldots, Z$. If `group = NULL`, all observations belong to group $z = 1$. |
| `y_id` | Observation index ($i = 1, \ldots, n_z$ within group $z$). |
| `y_name` | Variable label for `y_id`. |
| `rep_id` | Posterior draw index ($s = 1, \ldots, S$). |
| `rep_label` | Variable label for `rep_id`. |
| `value` | Sorted predicted probabilities $p_i^{(s)}$. |
| `ord` | Rank $k = g(i)$ of each sorted predicted probability within its group and draw. |
| `cep` | Conditional event probability $m_k$. |

## `ppc_calibration_overlay()` - one curve per posterior draw

`ppc_calibration_overlay()` draws one calibration curve per posterior draw.
This makes it easy to see both the central tendency and the draw-to-draw
variability of the calibration.

**How to read the plot:**

- **Calibrated**: curves cluster tightly around the diagonal.
- **Underprediction**: curves lie *above* the diagonal (observed rates exceed
  predicted rates).
- **Overprediction**: curves lie *below* the diagonal (predicted rates exceed
  observed rates).

```{r overlay-basic, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration_overlay(y = y, prep = prep_good) +
  labs(title = "Calibrated")
p2 <- ppc_calibration_overlay(y = y, prep = prep_bad) +
  labs(title = "Miscalibrated")

p1 | p2
```

### Trimming the x-axis with `x_range`

By default the x-axis spans the full [0, 1] interval (`x_range = "full"`).
When predictions are concentrated in a narrower range you can focus the plot
on the data by setting `x_range = "data"`:

```{r overlay-xrange, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration_overlay(y = y, prep = prep_good, x_range = "data") +
  labs(title = "Calibrated")
p2 <- ppc_calibration_overlay(y = y, prep = prep_bad, x_range = "data") +
  labs(title = "Miscalibrated")

p1 | p2
```

### Grouped data with `ppc_calibration_overlay_grouped()`

When observations belong to meaningful subgroups, pass a group vector to
`ppc_calibration_overlay_grouped()` to produce a faceted plot:

```{r overlay-grouped, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
gr <- c(replicate(N / 2, "Group-A"), replicate(N / 2, "Group-B"))

ppc_calibration_overlay_grouped(y = y, prep = prep_good, group = gr)
```

## `ppc_calibration()` — calibration curve with uncertainty bands

`ppc_calibration()` is the main plotting function. It displays the calibration
curve together with an uncertainty band. The meaning of the band depends on
the `interval` argument.

### Inputs: `prep` vs. `yrep`

The function accepts two types of input, suited to different modelling
workflows.

#### Using `y` and `prep` (matrix of predicted probabilities)

Supply:

- **`y`**: a length-$n$ vector of binary observations $y_i \in \{0, 1\}$.
- **`prep`**: an $S \times n$ matrix of predicted probabilities
  $p_i^{(s)} \in [0, 1]$ for each posterior draw $s$.

**Building the calibration curve:**

1. For each draw $s$, sort predicted probabilities in ascending order: $p_k^{(s)}$.
2. Reorder observations to match: $y_k^{(s)}$.
3. Compute the conditional event probability (CEP) $m_{\text{obs},k}^{(s)}$
   at each position $k$.
4. Average the CEPs across all $S$ draws to produce the plotted curve.

**Uncertainty bands:**

| `interval` | How it is computed | What it answers |
|---|---|---|
| `"confidence"` | Quantiles of $\{m_{\text{obs},k}^{(s)}\}_{s=1}^S$ | *Where would we expect our model's calibration curve to lie?* (**estimation uncertainty**) |
| `"consistency"` | Quantiles of CEPs computed from $\tilde{y}_k^{(s)} \sim \text{Bernoulli}(p_k^{(s)})$ | *Where would a perfectly calibrated model's curve lie?* (**posterior predictive consistency**) |

```{r setup-new-data}
# New simulated dataset for the ppc_calibration() examples
N <- 300
S <- 200
x <- runif(N, -2, 2)

p_true <- plogis(-0.2 + 1.1 * x)
y      <- rbinom(N, size = 1, prob = p_true)

prep <- matrix(
  plogis(rnorm(S * N, mean = rep(qlogis(p_true), each = S), sd = 0.35)),
  nrow = S, ncol = N
)

yrep <- matrix(
  rbinom(S * N, size = 1, prob = as.vector(prep)),
  nrow = S, ncol = N
)
```

```{r calib-prep-plot, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration(y, prep, interval = "consistency", show_qdots = FALSE)
p2 <- ppc_calibration(y, prep, interval = "confidence",  show_qdots = FALSE)

p1 | p2
```

#### Using `y` and `yrep` (matrix of posterior predictions)

Supply:

- **`y`**: a length-$n$ vector of binary observations $y_i \in \{0, 1\}$.
- **`yrep`**: an $S \times n$ matrix of posterior predictions
  $y_{\text{rep},i}^{(s)} \in \{0, 1\}$.

**Building the calibration curve:**

1. Estimate predicted probabilities as the column-wise posterior mean:
   $p_i = \frac{1}{S} \sum_{s=1}^S y_{\text{rep},i}^{(s)}$.
2. Sort probabilities in ascending order: $p_k$.
3. Reorder observations accordingly: $y_k$.
4. Compute the CEP $m_{\text{obs},k}$ which is the calibration curve.

**Uncertainty bands:**

| `interval` | How it is computed | What it answers |
|---|---|---|
| `"confidence"` | Bootstrap: for each replicate $b$, sample indices with replacement, sort by $p_j$, compute $m_{\text{obs},k}^{(b)}$ | *Where would we expect our model's curve to lie?* (**estimation uncertainty**) |
| `"consistency"` | For each draw $s$, compute the CEP of sorted posterior predictions $y_{\text{rep},k}^{(s)}$ | *Where would a calibrated model's curve lie?* (**posterior predictive consistency**) |

```{r calib-yrep-plot, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration(y, yrep, interval = "consistency", show_qdots = FALSE)
p2 <- ppc_calibration(y, yrep, interval = "confidence",  show_qdots = FALSE)

p1 | p2
```

### Additional arguments

#### `show_qdots` and `qdots_quantiles` — quantile dot plot

Setting `show_qdots = TRUE` overlays a quantile dot plot along the x-axis,
showing the marginal distribution of predicted probabilities. Each dot
represents an empirical quantile; the visual density is analogous to a
histogram but with low bias and higher variance. By default
`qdots_quantiles = 100` dots are displayed. See Säilynoja et al. (2025) for
methodological details.

```{r qdots, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration(y, prep, interval = "confidence", show_qdots = TRUE) +
  labs(title = "qdots_quantiles = 100")
p2 <- ppc_calibration(y, prep, interval = "confidence", show_qdots = TRUE,
                      qdots_quantiles = 200) +
  labs(title = "qdots_quantiles = 200")

p1 | p2
```

#### `x_range` - axis trimming

Setting `x_range = "data"` restricts the x-axis to the range of predicted
probabilities. This is useful when predictions are concentrated in a
sub-interval of [0, 1]:

```{r xrange, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration(y, prep, interval = "confidence", show_qdots = TRUE,
                      x_range = "data") +
  labs(title = 'x_range = "data"')
p2 <- ppc_calibration(y, prep, interval = "confidence", show_qdots = TRUE,
                      x_range = "full") +
  labs(title = 'x_range = "full"')

p1 | p2
```

#### `help_text` - interpretive caption

A short interpretive caption appears below the plot by default
(`help_text = TRUE`). Set `help_text = FALSE` to suppress it, for example
when embedding the plot in a report that provides its own caption:

```{r help-text, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration(y, prep, interval = "confidence", show_qdots = TRUE,
                      help_text = TRUE) +
  labs(title = "help_text = TRUE")
p2 <- ppc_calibration(y, prep, interval = "confidence", show_qdots = TRUE,
                      help_text = FALSE) +
  labs(title = "help_text = FALSE")

p1 | p2
```

## `ppc_calibration_interval_data()` - the interval data structure

`ppc_calibration_interval_data()` returns the numerical summaries that
underlie the uncertainty bands in `ppc_calibration()`. The `prob` argument
sets the desired coverage $(1 - \alpha)$, so the band spans from the
$\alpha/2$ to the $1 - \alpha/2$ quantile.

```{r interval-data, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
dat_interval <- ppc_calibration_interval_data(
  y        = y,
  prep     = prep,
  prob     = 0.95,
  interval = "consistency"
)

head(dat_interval, 5)
```

The returned columns are:

| Column | Description |
|---|---|
| `y_id` | Observation index ($i = 1, \ldots, n$). |
| `value` | Predicted probability $p_k$, averaged across posterior draws. |
| `cep` | Conditional event probability, averaged across posterior draws. |
| `lb` | Lower bound of the uncertainty interval. |
| `ub` | Upper bound of the uncertainty interval. |

---

## Real-world example: modelling roach infestation

The following example is drawn from Säilynoja et al. (2025) and uses the
`roaches` dataset from `rstanarm`. The data records the number of roaches
caught in traps across 264 apartments assigned to a treatment or control
condition. We compare two count-data models for the binary outcome "at least
one roach observed":

- A **negative binomial** model (fitted with `rstanarm::stan_glm()`).
- A **zero-inflated negative binomial** model (fitted with `brms::brm()`).

For a complete Bayesian workflow using this dataset see also Aki Vehtari's
case study
[*Roaches cross-validation model checking and comparison*](https://users.aalto.fi/~ave/casestudies/roaches/roaches.html).

```{r roaches-fit, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center", message=FALSE, warning=FALSE}
data(roaches)

roaches$sqrt_roach1 <- sqrt(roaches$roach1)
n <- length(roaches$y)

# Zero-inflated negative binomial (brms)
brm_glmzinb <- brms::brm(
  brms::bf(
    y ~ sqrt_roach1 + treatment + senior + offset(log(exposure2)),
    zi ~ sqrt_roach1 + treatment + senior + offset(log(exposure2))
  ),
  family = brms::zero_inflated_negbinomial(),
  data   = roaches,
  prior  = c(
    brms::prior(normal(0, 3), class = "b"),
    brms::prior(normal(0, 3), class = "b",         dpar = "zi"),
    brms::prior(normal(0, 3), class = "Intercept", dpar = "zi")
  ),
  seed    = SEED,
  refresh = 0
)

# Negative binomial (rstanarm)
stan_glmnb <- stan_glm(
  y ~ sqrt_roach1 + treatment + senior,
  offset          = log(exposure2),
  data            = roaches,
  family          = neg_binomial_2,
  prior           = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains          = 4,
  cores           = 1,
  seed            = SEED,
  refresh         = 0
)
```

```{r roaches-plot, fig.width=8, fig.height=3, fig.asp=NULL, out.width="90%", fig.align="center"}
pp <- pmin(brms::posterior_predict(stan_glmnb), 1)

p1 <- ppc_calibration(
  y             = pmin(roaches$y, 1),
  yrep          = pp,
  interval_type = "consistency",
  prob          = 0.95,
  x_range       = "full"
) +
  labs(title = "Negative binomial")

p2 <- ppc_calibration(
  y             = pmin(roaches$y, 1),
  yrep          = apply(posterior_predict(brm_glmzinb), 2, pmin, 1),
  interval_type = "consistency",
  prob          = 0.95,
  x_range       = "full"
) +
  labs(title = "Zero-inflated negative binomial")

p1 | p2
```

The calibration plots highlight a clear difference between the two models. The
**negative binomial** model is under-confident when predicting zero-roach
outcomes: its calibration curve falls outside the consistency band, indicating
a systematic mismatch between predicted and observed event rates. The
**zero-inflated** model's curve stays within the band and additionally covers
a wider range of predicted probabilities, reflecting both better calibration
and stronger discrimination between apartments with and without roaches.

## References

Kay, M. (2023). *ggdist: Visualizations of Distributions and Uncertainty*.
Zenodo. <https://doi.org/10.5281/zenodo.3879620>

Säilynoja, T., Johnson, A. R., Martin, O. A., & Vehtari, A. (2025).
Recommendations for visual predictive checks in Bayesian workflow.
<https://arxiv.org/abs/2503.01509>