---
title: "PPC Calibration"
author: "bayesplot team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{PPC Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="children/SETTINGS-knitr.txt"}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r pkgs}
library(brms)
library(dplyr)
library(rstanarm)
library("ggplot2")
library("patchwork")

SEED <- 840
set.seed(SEED)
```

## Overview

This vignette introduces the **PPC calibration** family of functions in
`bayesplot`, which assess the calibration of Bayesian models with binary
outcomes by examining the agreement between predicted probabilities and
observed event rates.

A model is *well-calibrated* when its predicted probabilities match empirical 
event frequencies. For example, among all observations where the model 
assigns a 30% probability, roughly 30% of events should actually occur. 
Deviations from this pattern signal systematic over- or underprediction.

This vignette explains the underlying methodology, walks through the core
plotting functions and their customization options, and showcases the use
of calibration plots in a real-world example. 

## Methodological Background

### Conditional Event Probabilities (CEP)

Following Dimitriadis, Gneiting, and Jordan (2021), `bayesplot` does not work
directly with raw binary outcomes. Instead, it estimates the **Conditional
Event Probability (CEP)**: the true underlying probability that an event
occurs, given that the model assigned a specific predicted probability $p$.

Formally, if the model predicts $p$ for an observation, the CEP answers:
*"Among all observations that received this prediction, what fraction actually
had the event?"*

A perfectly calibrated model has a CEP function equal to the identity, that
is, CEP($p$) = $p$ for all $p$. When plotted, this produces a diagonal reference
line. Deviations from the diagonal reveal miscalibration.

### Estimating the CEP: the PAV Algorithm

To estimate the CEP from data, `bayesplot` uses the **Pool Adjacent
Violators (PAV)** algorithm (Ayer et al., 1955). For more details see
Dimitriadis et al. (2021) and Säilynoja et al. (2025). The PAV algorithm solves
an isotonic regression problem: it finds the monotone non-decreasing step
function that best fits the binary outcomes as a function of the ordered
predicted probabilities.

The estimation procedure has three steps which are described below and
illustrated with a toy example consisting of 8 observations with their
predicted probabilities $p_i$ and binary outcomes $y_i$ for $i = 1, \ldots, 8$.

#### Step 1: Sort by predicted probability

Order observations by their predicted probability.

```{r toy-data}
df_toy <- data.frame(
  p = c(0.05, 0.10, 0.20, 0.30, 0.45, 0.60, 0.75, 0.90),
  y = c(0,    0,    1,    0,    1,    1,    0,    1)
)
df_toy
```

#### Step 2: Fit a monotone step function with PAV

Apply the PAV algorithm to the ordered outcomes. The result is a piecewise-
constant, non-decreasing estimate of the CEP at each predicted
probability $p_i$.

```{r pav-estimate}
df_toy <- df_toy |>
  dplyr::mutate(cep = round(stats::isoreg(df_toy$y)$yf, 2))
df_toy
```

#### Step 3: Plot CEP against predicted probability

Plot the estimated step function with predicted probabilities $p_i$ on the
x-axis and CEP on the y-axis. The diagonal reference line marks perfect
calibration.

```{r toy-plot, out.width = "60%", fig.align = "center", echo = FALSE}
S <- 5
ppc_calibration(
  y = df_toy$y, prep = t(replicate(S, df_toy$p)), show_qdots = FALSE
  )
```

In a fitted Bayesian model the predicted probabilities are the posterior
predictions and the CEP is estimated for each posterior draw separately. In the
following code block we adjust the toy example by simulating $S=5$ draws of
predicted probabilities for each observation.

```{r toy-data2}
S <- 5

df_toy_draws <- list(
  p = t(replicate(S, pmin(pmax(df_toy$p + rnorm(length(df_toy$p), 0, 0.08), 0), 1))),
  y = c(0, 0, 1, 0, 1, 1, 0, 1)
)
```

The calibration curve can then be plotted for each draw using
`ppc_calibration_overlay()`. The resulting plot shows the draw-to-draw
variability in the calibration curve.

```{r toy-plot2, out.width = "60%", fig.align = "center"}
ppc_calibration_overlay(y = df_toy_draws$y, prep = df_toy_draws$p, alpha = 0.9)
```

Alternatively, `ppc_calibration()` can be used to summarize the draw-specific
curves into a single calibration curve with an uncertainty band. The next
section explains how the uncertainty intervals are constructed.

```{r toy-plot3, out.width = "60%", fig.align = "center"}
ppc_calibration(y = df_toy_draws$y, prep = df_toy_draws$p, show_qdots = FALSE)
```

### Uncertainty Intervals {#uncertainty-intervals}

`ppc_calibration()` summarizes the draw-specific curves into a single
calibration curve with an uncertainty band. The plotting function supports two
strategies for propagating uncertainty via the `interval` argument.

#### Estimation uncertainty (`interval = "confidence"`): 

The default setting `interval = "confidence"` reflects uncertainty in the CEP
induced by variation across posterior draws:

1. For each posterior draw $s$, sort observations $y_i$ by posterior predictions
   $p_i^{(s)}$.
2. Estimate for each posterior draw a CEP curve.
3. Take pointwise quantiles of the draw-specific CEP values across all $S$
   draws (e.g., 2.5% and 97.5% for `prob = 0.95`) to form the ribbon.

The central step curve is the pointwise posterior mean of the draw-specific
curves.

In the `ppc_calibration()` plot, this interval type is labeled `95%-CI (ptw.)`,
indicating a pointwise confidence interval for the CEP curve whose bounds are
the 2.5% and 97.5% quantiles of the draw-specific CEP curves. The probability
can be adjusted with the `prob` argument.

#### Posterior predictive consistency (`interval = "consistency"`)

The setting `interval = "consistency"` answers the question: *If the model is
correctly specified, where would we expect the calibration curve to fall?*
The band is constructed as follows:

1. For each draw $s$ and its sorted posterior predictions $p_i^{(s)}$,
   simulate replicated outcomes
   $\tilde{y}_i^{(s)} \sim \mathrm{Bernoulli}(p_i^{(s)})$.
2. Estimate for each replicated outcome a CEP curve.
3. Take pointwise quantiles of the CEP values across all $S$ draws to
   form the ribbon.

The central curve is still estimated from the *observed* outcomes. If the
observed curve falls outside the consistency band, the model is
self-**in**consistent, that is, its predictions do not match the data-generating
distribution it implies.

In the `ppc_calibration()` plot, this interval type is labeled `95%-CsI (ptw.)`,
indicating a pointwise consistency interval for the CEP curve whose bounds are
the 2.5% and 97.5% quantiles of the draw-specific CEP curves. The probability
can be adjusted with the `prob` argument.

### Calibrated vs. Miscalibrated: an Illustration

The following figure contrasts a calibrated and a miscalibrated model fitted to
simulated data, and demonstrates the interpretation of the diagonal reference
line introduced above.

We simulate $n=600$ observations of a single predictor $x \sim N(0,1)$ and 
generate two binary outcomes. 

+ The **calibrated** outcome follows a logistic model linear in $x$, so the 
same linear model fitted by `brm()` is correctly specified and its predictions 
should align with the observed event rates. 

+ The **miscalibrated** outcome is generated from a model that also includes a 
quadratic term $0.9 \times x^2$, which the fitted linear model cannot capture; 
this systematic misspecification causes the predicted probabilities to diverge 
from the true event rates. 

Both models share the same weakly informative priors and are fitted with a 
single chain for illustration purposes.

```{r simulate-data, warning = FALSE, message = FALSE}
n <- 600

x <- rnorm(n)
y <- rbinom(n, size = 1, prob = plogis(-0.5 + 1.2 * x))
y_mis <- rbinom(n, size = 1, prob = plogis(-0.5 + 1.2 * x + 0.9 * x^2))

df <- data.frame(y = y, x = x)
df_mis <- data.frame(y = y_mis, x = x)

fit_model <- function(df, seed) {
  brm(
    formula = y ~ x,
    data = df,
    family = bernoulli(link = "logit"),
    prior = c(
      prior(normal(0, 2.5), class = "b"),
      prior(normal(0, 5), class = "Intercept")
    ),
    chains = 1,
    seed = seed
  )
}

fit_calib <- fit_model(df, SEED)
fit_miscalib <- fit_model(df_mis, SEED)

prep <- posterior_epred(fit_calib)
prep_mis <- posterior_epred(fit_miscalib)

yrep <- posterior_predict(fit_calib)
yrep_mis <- posterior_predict(fit_miscalib)
```

While the curve in the calibrated model (left) tracks the diagonal closely,
the miscalibrated model (right) shows a systematic deviation from the diagonal
reference line.

```{r calib-vs-miscalib, fig.width = 7, fig.height = 3, fig.asp = NULL, out.width = "90%", fig.align = "center", echo = FALSE}
p1 <- ppc_calibration(y = y, yrep = yrep, show_qdots = FALSE) +
  labs(title = "Calibrated model")

p2 <- ppc_calibration(y = y_mis, yrep = yrep_mis, show_qdots = FALSE) +
  labs(title = "Miscalibrated model")

p1 + p2
```

## Overview of ppc-calibration functions and customization options
In the following an overview of the core `ppc_calibration` functions is 
provided, along with explanations of their input arguments and customization 
options. 

### `ppc_calibration_data()` — the underlying data structure

`ppc_calibration_data()` computes the data that all calibration plots are built
on. Understanding its output helps to work with or extend the visualisations or
to build custom calibration plots.

```{r ppc-calibration-data}
dat <- ppc_calibration_data(y = y, prep = prep)

head(dat, 5)
```

`ppc_calibration_data` returns a data frame with the following columns:

| Column | Description |
|---|---|
| `group` | Group label; if `group = NULL`, all observations belong to one group. (factor or double) |
| `y_id` | Observation index ($i = 1, \ldots, n_z$ within group $z$). (integer) |
| `y_name` | Variable label for `y_id`. (factor) |
| `rep_id` | Posterior draw index ($s = 1, \ldots, S$). (integer) |
| `rep_label` | Variable label for `rep_id`. (factor) |
| `value` | Sorted predicted probabilities $p_i^{(s)}$. (double) |
| `ord` | Rank of each sorted predicted probability within its group and draw. (integer) |
| `cep` | Conditional event probability $cep_i^{(s)}$. (double) |

### `ppc_calibration_overlay()` — one curve per posterior draw

`ppc_calibration_overlay()` draws one calibration curve per posterior draw,
making it a useful tool for exploring the variability in calibration across
posterior draws.

In the following we use the simulated data from the previous section to compare 
the draw-specific calibration curves of the calibrated and miscalibrated models.
The plot shows that the curves from the calibrated model (left) cluster around 
the diagonal reference line, while those from the miscalibrated model (right) 
deviate from it for lower predicted probabilities.

```{r overlay-basic, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration_overlay(y = y, prep = prep) +
  labs(title = "Calibrated")

p2 <- ppc_calibration_overlay(y = y, prep = prep_mis) +
  labs(title = "Miscalibrated")

p1 + p2
```

#### Grouped data

When observations belong to subgroups, `ppc_calibration_overlay_grouped()` 
produces a faceted plot. 

For illustration, we define for our simulated data a grouping variable by 
assigning the first half of the simulated observations to group A and the 
second half to group B.

```{r overlay-grouped, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}

group <- rep(c("A", "B"), each = n / 2)

ppc_calibration_overlay_grouped(y = y, prep = prep, group = group)
```

### `ppc_calibration()` — calibration curve with uncertainty bands

`ppc_calibration()` is the main plotting function. It displays the calibration
curve together with an uncertainty band. The meaning of the band depends on
the `interval` argument (as discussed above).

#### Expected input arguments

`ppc_calibration()` expects the following input arguments:

- **Core data**
  - **`y`**: numeric vector of observed outcomes of length `N`, no `NA`, with
    values in `[0, 1]` (typically binary `0/1`).
  - Exactly one of:
    - **`prep`**: numeric `S x N` matrix of predicted probabilities in `[0, 1]`,
      no `NA`, where `S` refers to the number of posterior draws.
    - **`yrep`**: numeric `S x N` matrix of posterior predictive draws,
      no `NA` (typically binary draws), where `S` refers to the number of
      posterior draws.
  - `ncol(prep)` or `ncol(yrep)` must equal `length(y)`.

- **Interval controls**
  - **`prob`**: a single value strictly between `0` and `1`.
  - **`interval`**: one of `"confidence"` or `"consistency"`.
  - **`B`**: if `interval = "consistency"`: positive integer indicating number
    of bootstrap samples.

- **Plot controls**
  - **`help_text`**: whether label (e.g., `95%-CI (ptw.)`) is shown.
  - **`show_mean`**: logical; whether the mean calibration curve is shown.
  - **`show_qdots`**: logical; whether the marginal distribution of predicted
    probabilities is shown as a quantile dot plot along the x-axis.
  - **`qdots_quantiles`**: positive integer; number of quantiles to display in
    the quantile dot plot (only if `show_qdots = TRUE`).

- **Styling**
  - **`linewidth`**, **`alpha`**: passed to ggplot geoms.

Because the function signature is
`ppc_calibration(y, prep = NULL, yrep = NULL, ...)`,
an unnamed second matrix argument is interpreted as `prep`. Use `yrep = ...`
explicitly when supplying posterior predictive draws.

#### `interval` and `prob` — uncertainty intervals

The `interval` argument controls the type of uncertainty interval displayed in
the plot. The `prob` argument controls the width of the interval (e.g., `0.95`
for a 95% interval). See the section on 
[Uncertainty Intervals](#uncertainty-intervals) above for details.

The following plot shows the calibration curves for the simulated calibrated and miscalibrated models, with both confidence intervals (top row) and consistency intervals (bottom row). 
```{r intervals, fig.width=7, fig.height=7, fig.asp=NULL, out.width="80%", fig.align="center"}

p1 <- ppc_calibration(y = y, yrep = yrep) +
  labs(title = "Calibrated model")

p2 <- ppc_calibration(y = y, yrep = yrep_mis) +
  labs(title = "Miscalibrated model")

p3 <- ppc_calibration(y = y, yrep = yrep, interval = "consistency")

p4 <- ppc_calibration(y = y, yrep = yrep_mis, interval = "consistency")

p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
```

#### `show_qdots` and `qdots_quantiles` — quantile dot plot

Setting `show_qdots = TRUE` overlays a quantile dot plot along the x-axis,
showing the marginal distribution of predicted probabilities. Each dot
represents an empirical quantile. By default `qdots_quantiles = 100` dots are 
displayed. See Säilynoja et al. (2025) for methodological details.

The following plot shows the calibration curves for the simulated calibrated and miscalibrated models, with quantile dot plots along the x-axis showing the 
distribution of predicted probabilities. The top row shows 100 quantiles, 
while the bottom row shows 300 quantiles.

```{r qdots, fig.width=7, fig.height=7, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration(y = y, yrep = yrep) +
  labs(title = "Calibrated model \n # qdots = 100")

p2 <- ppc_calibration(y = y, yrep = yrep_mis) +
  labs(title = "Miscalibrated model \n # qdots = 100")

p3 <- ppc_calibration(y = y, yrep = yrep, qdots_quantiles = 300) +
  labs(title = "# qdots = 300")

p4 <- ppc_calibration(y = y, yrep = yrep_mis, qdots_quantiles = 300) +
  labs(title = "# qdots = 300")

p1 + p2 + p3 + p4 + plot_layout(ncol = 2)
```

#### `help_text` — interpretive caption

A short caption appears in the plot by default (`help_text = TRUE`) which
provides information about the interval type and probability. This is intended
to help users interpret the plot correctly. It can be suppressed by setting
`help_text = FALSE`.

The following plot shows the calibration curve for the simulated calibrated 
model with and without the interpretive caption.

```{r help-text, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center"}
p1 <- ppc_calibration(y = y, yrep = yrep, help_text = TRUE)

p2 <- ppc_calibration(y = y, yrep = yrep, help_text = FALSE)

p1 | p2
```

### `ppc_loo_calibration()` — LOO-adjusted calibration curve

`ppc_loo_calibration()` is a variant of `ppc_calibration()` that incorporates
importance weights from a leave-one-out (LOO) analysis. Weights can be provided
directly via the `lw` argument; alternatively, a `psis_object` can be passed 
and the weights will be computed internally. 

As with `ppc_calibration()`, a grouped version `ppc_loo_calibration_grouped()` 
is available that produces a faceted plot when a grouping variable is supplied.

## Real-World Example: Modelling Roach Infestation

The following example is drawn from Säilynoja et al. (2025) and uses the
`roaches` dataset from `rstanarm`. The data records the number of roaches
caught in traps across 264 apartments assigned to a treatment or control
condition. We compare two count-data models for the binary outcome "at least
one roach observed":

- A **negative binomial** model (fitted with `rstanarm::stan_glm()`).
- A **zero-inflated negative binomial** model (fitted with `brms::brm()`).

For a complete Bayesian workflow using this dataset see also Aki Vehtari's
case study [*Roaches cross-validation model checking and comparison*](https://users.aalto.fi/~ave/casestudies/roaches/roaches.html).

```{r roaches-fit, fig.width=7, fig.height=3, fig.asp=NULL, out.width="80%", fig.align="center", message=FALSE, warning=FALSE, results = "hide"}
SEED <- 840
data(roaches, package = "rstanarm")

roaches$sqrt_roach1 <- sqrt(roaches$roach1)
n <- length(roaches$y)

brm_glmzinb <- brms::brm(
  formula = brms::bf(
    y ~ sqrt_roach1 + treatment + senior + offset(log(exposure2)),
    zi ~ sqrt_roach1 + treatment + senior + offset(log(exposure2))
  ),
  family = brms::zero_inflated_negbinomial(),
  data = roaches,
  prior = c(
    brms::prior(normal(0, 3), class = "b"),
    brms::prior(normal(0, 3), class = "b", dpar = "zi"),
    brms::prior(normal(0, 3), class = "Intercept", dpar = "zi")
  ),
  seed = SEED,
  refresh = 0,
  silent = 2
)

stan_glmnb <- rstanarm::stan_glm(
  formula = y ~ sqrt_roach1 + treatment + senior,
  offset = log(exposure2),
  data = roaches,
  family = neg_binomial_2,
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains = 4,
  cores = 1,
  seed = SEED,
  refresh = 0
)
```

### Confidence intervals for estimation uncertainty

Where do we expect our model's calibration curve to lie?

```{r roaches-plot-CI, fig.width=7, fig.height=3, fig.asp=NULL, out.width="90%", fig.align="center"}
pp_nb <- pmin(brms::posterior_predict(stan_glmnb), 1)
pp_zinb <- apply(brms::posterior_predict(brm_glmzinb), 2, pmin, 1)

p1 <- ppc_calibration(
  y = pmin(roaches$y, 1),
  yrep = pp_nb,
  interval = "confidence",
  prob = 0.95
) +
  labs(title = "Negative binomial")

p2 <- ppc_calibration(
  y = pmin(roaches$y, 1),
  yrep = pp_zinb,
  interval = "confidence",
  prob = 0.95
) +
  labs(title = "Zero-inflated negative binomial")

p1 + p2
```

### Consistency intervals for model checking

Where would the calibration curve of a calibrated model lie?

```{r roaches-plot-CsI, fig.width=7, fig.height=3, fig.asp=NULL, out.width="90%", fig.align="center"}
p1 <- ppc_calibration(
  y = pmin(roaches$y, 1),
  yrep = pp_nb,
  interval = "consistency",
  prob = 0.95
) +
  labs(title = "Negative binomial")

p2 <- ppc_calibration(
  y = pmin(roaches$y, 1),
  yrep = pp_zinb,
  interval = "consistency",
  prob = 0.95
) +
  labs(title = "Zero-inflated negative binomial")

p1 + p2
```

The `ppc_calibration()` plot indicates that the **negative binomial** model is
underconfident when predicting zero-roach outcomes: its calibration curve falls
outside the consistency band, revealing a mismatch between predicted and
observed event rates.

The **zero-inflated** model's curve stays within the band and spans a wider
range of predicted probabilities, reflecting both better calibration and
stronger discrimination between apartments with and without roaches.

### LOO calibration

`ppc_loo_calibration()` uses the `brm_glmzinb` object fitted above. The PSIS
object is computed from it and then passed alongside the standard arguments.
The comparison below shows the in-sample calibration curve (left) and the
LOO-adjusted curve (right) for the negative binomial model.

```{r calibration-loo-plot, fig.width=7, fig.height=3, fig.asp=NULL, out.width="90%", fig.align="center"}
brm_glmzinb <- brms::add_criterion(brm_glmzinb, criterion = "loo")
fit_glmzinb_loo <- brms::loo(brm_glmzinb, save_psis = TRUE)

p1 <- ppc_calibration(
  y = pmin(roaches$y, 1),
  yrep = pp_nb,
  interval = "consistency",
  prob = 0.95
) +
  labs(title = "Negative binomial")

p2 <- ppc_loo_calibration(
  y = pmin(roaches$y, 1),
  yrep = pp_nb,
  psis_object = fit_glmzinb_loo$psis_object,
  interval = "consistency",
  prob = 0.95
) +
  labs(title = "Negative binomial (LOO)")

p1 + p2
```

## References

- Ayer, M., Brunk, H. D., Ewing, G. M., Reid, W. T., & Silverman, E. (1955).
  An empirical distribution function for sampling with incomplete information.
  *Annals of Mathematical Statistics*, 26(4), 641–647.

- Dimitriadis, T., Gneiting, T., & Jordan, A. I. (2021).
  Stable reliability diagrams for probabilistic classifiers.
  *Proceedings of the National Academy of Sciences*, 118(8).
  https://doi.org/10.1073/pnas.2016191118

- Säilynoja, T., Johnson, A. R., Martin, O. A., & Vehtari, A. (2025).
  Recommendations for visual predictive checks in Bayesian workflow.
  (Preprint). *arXiv*. https://doi.org/10.48550/arXiv.2503.01509