---
title: "Graphical posterior predictive checks using the bayesplot package"
author: "Jonah Gabry"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Graphical posterior predictive checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, settings, include=FALSE}
library("bayesplot")
knitr::opts_chunk$set(
  dev = "pdf",
  fig.align = "center",
  fig.width = 4,
  fig.height = 4,
  comment = NA
)
```

## Overview

The __bayesplot__ package provides various plotting functions for 
_graphical posterior predictive checking_, that is, creating graphical displays
comparing observed data to simulated data from the posterior predictive
distribution.

The idea behind posterior predictive checking is simple: if a model is a
good fit then we should be able to use it to generate data that looks a lot like
the data we observed.

#### Posterior predictive distribution
To generate the data used for posterior predictive checks (PPCs) we simulate
from the _posterior predictive distribution_ The posterior predictive 
distribution is the distribution of the outcome variable implied by a model 
after using the observed data $y$ (a vector of $N$ outcome values) to update our
beliefs about unknown model parameters $\theta$, $$p(\tilde{y} \,|\, y) = \int
p(\tilde{y} \,|\, \theta) \, p(\theta \,|\, y) \, d\theta.$$
Typically we will also condition on $X$ (a matrix of predictor variables).

For each draw (simulation) $s = 1, \ldots, S$ of the parameters from the 
posterior distribution, $\theta_s \sim p(\theta \,|\, y)$, we draw an entire
vector of $N$ outcomes $y^{rep}_s$ from the posterior predictive distribution
by simulating from the data model conditional on parameters $\theta_s$.
The result is an $S \times N$ matrix of draws $y^{rep}$.

When simulating from the posterior predictive distribution we can use either the
same values of the predictors $X$ that we used when fitting the model or new 
observations of those predictors. When we use the same values of $X$ we denote 
the resulting simulations by $y^{rep}$, as they can be thought of as 
replications of the outcome $y$ rather than predictions for future observations 
($\tilde{y}$ using predictors $\tilde{X}$). This corresponds to the notation 
from Gelman et. al. (2013) and is the notation used throughout the package
documentation.


## Graphical posterior predictive checks

Using the replicated datasets drawn from the posterior predictive
distribution, the functions in the __bayesplot__ package create various
graphical displays comparing the observed data $y$ to the replications.
The names of the __bayesplot__ plotting functions for posterior predictive
checking all have the prefix `ppc_`. 

To demonstrate the various PPCs that can be created with the __bayesplot__ 
package we'll use an example of comparing Poisson and Negative binomial
regression models from the
[**rstanarm**](https://cran.r-project.org/web/packages/rstanarm/) package
vignette [_stan_glm: GLMs for Count
Data_](https://cran.r-project.org/web/packages/rstanarm/vignettes/count.html).

> We want to make inferences about the efficacy of a certain pest management system at reducing the number of roaches in urban apartments. [...] 
The regression predictors for the model are the pre-treatment number of roaches `roach1`, the treatment indicator `treatment`, and a variable `senior` indicating whether the apartment is in a building restricted to elderly residents. Because the number of days for which the roach traps were used is not the same for all apartments in the sample, we include it as an exposure [...]. 

First we fit a Poisson regression model with outcome variable `y` representing 
the roach count in each apartment at the end of the experiment.

```{r, roaches-model, results="hide", message=FALSE,warning=FALSE}
library("rstanarm")
data(roaches) # see help("rstanarm-datasets")
head(roaches)

roaches$roach1 <- roaches$roach1 / 100 # pre-treatment number of roaches (in 100s)
fit_poisson <- stan_glm(
  y ~ roach1 + treatment + senior, 
  data = roaches,
  offset = log(exposure2), 
  family = poisson(link = "log"), 
  prior = normal(0,2.5), 
  prior_intercept = normal(0,5),
  chains = 4, 
  iter = 2000,
  seed = 1111
)
```

```{r, print}
print(fit_poisson)
```

In order to use the PPC functions from the __bayesplot__ package we need
a matrix of draws from the posterior predictive distribution. Since we fit 
the model using __rstanarm__ we can use its `posterior_predict` function:

```{r, posterior_predict}
yrep <- posterior_predict(fit_poisson)
dim(yrep)
```

Each row of this matrix is a draw from the posterior predictive distribution, 
i.e. a vector with `r ncol(yrep)` elements, one for each of the data points in
`y`.

### Histograms and density estimates

The first PPC we'll look at is a comparison of the distribution of `y` and the
distributions of some of the simulated datasets (rows) in the `yrep` matrix.

```{r ppc_dens_overlay}
library("bayesplot")

y <- roaches$y
ppc_dens_overlay(y, yrep[1:50, ])
```

In the plot above, the dark line is the distribution of the observed outcomes 
`y` and each of the lighter lines is the kernel density estimate of one of the
replications of `y` from the posterior predictive distribution. This 
plot makes it easy to see that this model fails to account for large 
proportion of zeros in `y`. That is, the model predicts fewer zeros than 
were actually observed.

We could see the same thing by looking at separate histograms of `y` and some 
of the `yrep` datasets using the `ppc_hist` function:

```{r ppc_hist}
ppc_hist(y, yrep[1:8, ])
```


### Scatterplots

Write this section.

### Distributions of test statistics

Write this section.

### Distributions of residuals

Write this section.

### Checks for time series data

Write this section.


## Providing an interface to bayesplot PPCs from another package

The __bayesplot__ package provides the S3 generic function `pp_check`. Authors of
R packages for Bayesian inference are encouraged to define methods for the
fitted model objects created by their packages. This will hopefully be
convenient for both users and developers and contribute to the use of the same
naming conventions across many of the R packages for Bayesian data analysis.


To provide an provide interface to __bayesplot__ from your package, you can very 
easily define a `pp_check` method (or multiple `pp_check` methods) for the
fitted model objects created by your package. All a `pp_check` method needs to
do is provide the `y` vector and `yrep` matrix arguments to the various plotting
functions included in __bayesplot__.

Here is an example for how to define a simple `pp_check` method in a package
that creates fitted model objects of class `"foo"`. We will define a method
`pp_check.foo` that extracts the data `y` and the draws from the posterior
predictive distribution `yrep` from an object of class `"foo"` and then calls 
one of the plotting functions from __bayesplot__.

Suppose that objects of class `"foo"` are lists with named components, two of 
which are `y` and `yrep`. Here's a simple method `pp_check.foo` that offers the
user the option of two different plots:

```{r, pp_check.foo}
pp_check.foo <- function(object, ..., type = c("multiple", "overlaid")) {
  y <- object[["y"]]
  yrep <- object[["yrep"]]
  switch(
    match.arg(type),
    multiple = ppc_hist(y, yrep[1:min(8, nrow(yrep)),, drop = FALSE]),
    overlaid = ppc_dens_overlay(y, yrep)
  )
}

print(methods("pp_check"))
```

To try out `pp_check.foo` we can just make a list with `y` and `yrep` components
and give it class `foo`:

```{r, foo-object}
x <- list(y = rnorm(50), yrep = matrix(rnorm(5000), nrow = 100, ncol = 50))
class(x) <- "foo"
```
```{r, pp_check-1, eval=FALSE}
bayesplot::pp_check(x)
```
```{r, print-1, echo=FALSE}
gg <- bayesplot::pp_check(x)
suppressMessages(print(gg))
```
```{r, pp_check-2, eval=FALSE}
bayesplot::pp_check(x, type = "overlaid")
```
```{r, print2, echo=FALSE}
bayesplot::pp_check(x, type = "overlaid")
```


## References

Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., and
Rubin, D. B. (2013). *Bayesian Data Analysis*. Chapman & Hall/CRC Press, London,
third edition.

Stan Development Team. (2016). *Stan Modeling Language Users
Guide and Reference Manual*. http://mc-stan.org/documentation/
