---
title: "Model checking using 'ppc_pit_ecdf' and 'ppc_loo_pit_ecdf'"
author: "Florence Bockting"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
params:
  EVAL: TRUE #!r identical(Sys.getenv("NOT_CRAN"), "true")
vignette: >
  %\VignetteIndexEntry{loo-pit-correlated-tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, child="../children/SETTINGS-knitr.txt"}
```

```{r pkgs, include=FALSE}
library(ggplot2)
library(patchwork)
library(loo)

bayesplot_theme_set()

set.seed(2026)

# Reusable figure chunk options (use via opts.label = "...")
knitr::opts_template$set(
  fig_wide = list(
    fig.width = 7,
    fig.height = 3,
    fig.asp = NULL,
    out.width = "80%",
    fig.align = "center",
    warning = FALSE,
    message = FALSE
  ),
  fig_square = list(
    out.width = "80%",
    fig.align = "center",
    warning = FALSE,
    echo = FALSE
  ),
  fig_grouped = list(
    fig.width = 7,
    fig.height = 3.5,
    fig.asp = NULL,
    out.width = "80%",
    fig.align = "center",
    warning = FALSE,
    message = FALSE
  )
)
```

## Setup
```{r, eval=FALSE}
library(bayesplot)
library(ggplot2)
library(patchwork)
library(loo)

set.seed(2026)
```

## Introduction
This vignette focuses on pointwise predictive checks based on the probability
integral transform (PIT). The PIT is defined as the model’s univariate
predictive cumulative distribution function (CDF) evaluated at the
observed data. For a well-calibrated model, these transformations result in 
values that should be approximately uniformly distributed between 0 and 1
(Tesso & Vehtari, [2026](#Tesso2026); Säilynoja et al., [2022](#Säilynoja2022)).

In `bayesplot`, PIT-based checks are implemented in the `ppc_pit_ecdf()` and
`ppc_loo_pit_ecdf()` functions, which plot the empirical CDF of the PIT values
against the expected uniform distribution. The `ppc_pit_ecdf_grouped()` function
extends this functionality to allow for grouping by covariates.

**New Correlation-Aware Method**

Following the work of Tesso & Vehtari ([2026](#Tesso2026)), `bayesplot` now
offers a new method for evaluating the uniformity of PIT values that accounts
for their correlation structure. This method is implemented via the argument
`method = "correlated"` and is recommended for most applications. The
previous approach is retained via `method = "independent"` for backward
compatibility and is the current default to simplfy transition for the user.

This vignette focuses specifically on the changes introduced by the new
correlation-aware method. For background information on graphical uniformity
tests using PIT, see Säilynoja et al. ([2022](#Säilynoja2022)). For a more
general discussion on the use of Leave-One-Out Cross-Validation (LOO-CV), see
Vehtari et al. ([2017](#Vehtari2017), [2024](#Vehtari2024)), among others.

## Reading the plots for different (mis)calibration scenarios
The shape of the ECDF curve provides direct insight into *how* a predictive
distribution is miscalibrated. To illustrate this, the following examples
utilize simulated scenarios where "observed" values (`y`) are drawn from a
`normal(0, sd)` distribution, while "replicated" values (`yrep`) are generated
from a non-central t-distribution. By varying the degrees of freedom (`df`) and
non-centrality parameter (`ncp`), we can simulate and visualize several 
distinct types of miscalibration.

```{r, echo=FALSE}
plot_scenarios <- function(
  df = NULL, ncp = NULL, sd = 1, n = 300, S = 100, seed = 2026,
  xlim = c(-5, 5)
  ) {
  set.seed(seed)
  y <- rnorm(n, mean = 0, sd = sd)
  yrep <- matrix(rt(n * S, df = df, ncp = ncp), nrow = S, ncol = n)

  yrep_df <- data.frame(
    value = as.vector(t(yrep)),
    draw  = rep(seq_len(nrow(yrep)), each = ncol(yrep))
  )

  p1 <- ggplot(
    yrep_df, 
    aes(x = .data$value, group = .data$draw, color = "yrep")
    ) +
    geom_density(linewidth = 0.5, alpha = 0.3) +
    geom_density(
      data = data.frame(value = y),
      aes(x = .data$value, color = "y"),
      linewidth = 1,
      inherit.aes = FALSE
    ) +
    scale_color_manual(
      values = c(yrep = "lightblue", y = "darkblue"),
      breaks = c("yrep", "y"),
      name = NULL
    ) +
    xlim(c(xlim[1], xlim[2])) +
    yaxis_ticks(FALSE) +
    xlab("y / yrep") +
    bayesplot_theme_get() +
    theme(
      legend.position = "top",
      legend.direction = "horizontal"
    )
  p2 <- ppc_pit_ecdf(y = y, yrep = yrep, method = "correlated", prob = 0.95)
  p3 <- ppc_pit_ecdf(y = y, yrep = yrep, method = "correlated",
    plot_diff = TRUE, prob = 0.95)

  bayesplot_grid(
    p1, p2, p3,
    grid_args = list(ncol = 3, widths = c(1.1, 1.1, 1.1))
    )
}

```

### Calibrated model
We begin by simulating a well-calibrated scenario where the predictive
distribution (`ypred`) closely matches the data-generating process (`y`). In
this case, the PIT values should be approximately uniform, and the ECDF should
closely follow the identity (diagonal) line.

$$
\begin{align*}
y &\sim \text{Normal}(0, 1) \\
y_{rep} &\sim \text{Student}_{\nu = 100}(0, 1) \quad (\approx \text{Normal}(0, 1))
\end{align*}
$$

```{r calibrated, opts.label="fig_wide", echo = FALSE}
plot_scenarios(df = 100, ncp = 0)
```

### Predictive distribution too wide (overdispersed)
When the predictive distribution is too wide, the model overestimates the
probability mass in the tails. As a result, observed values tend to fall near
the *centre* of the predictive distribution more often than expected. 
This leads to an `S`-shaped ECDF curve that typically falls below the diagonal 
line in the lower tail and rises above it in the upper tail.

$$
\begin{aligned}
y &\sim \text{Normal}(0, 1) \\
y_{rep} &\sim \text{Student}_{\nu = 1}(0, 1) \quad (= \text{Cauchy}(0, 1))
\end{aligned}
$$

```{r too-wide, opts.label="fig_wide", echo = FALSE}
plot_scenarios(df = 1, ncp = 0, xlim = c(-7, 7))
```

### Predictive distribution too narrow (underdispersed)
When the predictive distribution is too narrow, the model underestimates the
potential for extreme values, causing observations to fall frequently in the
tails. This results in a reverse `S`-shaped ECDF curve that typically lies 
above the diagonal line in the lower tail and below it in the upper tail.

$$
\begin{aligned}
y &\sim \text{Normal}(0, 2) \\
y_{rep} &\sim \text{Student}_{\nu = 100}(0, 1) \quad (\approx \text{Normal}(0, 1))
\end{aligned}
$$

```{r too-narrow, opts.label="fig_wide", echo = FALSE}
plot_scenarios(df = 100, ncp = 0, sd = 2)
```

### Systematically biased predictions (location shift)
When the model's predictions are systematically shifted relative to the data,
observations tend to fall consistently in one tail of the predictive
distribution. This results in an ECDF curve that "bows" aways from the identity
line:

+ Underestimation (Positive Bias): If the model predicts values that are
systematically too low, the PIT values will cluster near 1. The ECDF curve will
stay below the diagonal line.
+ Overestimation (Negative Bias): If the model predicts values that are
systematically too high, the PIT values will cluster near 0. The ECDF curve will
stay above the diagonal line.

$$
\begin{aligned}
y &\sim \text{Normal}(0, 1) \\
y_{rep} &\sim \text{Student}_{\nu = 100}(-1, 1)
\end{aligned}
$$

```{r biased, opts.label="fig_wide", echo = FALSE}
plot_scenarios(df = 100, ncp = -1)
```

## The `method` argument
### `method = "independent"` (superseded)
When `method = "independent"` is selected, simultaneous confidence bands for
the ECDF are constructed under the assumption that the PIT values are both
independent and uniform (Säilynoja et al., [2022](#Säilynoja2022)). However, if
this independence assumption is violated, the resulting bands can be too wide,
which reduces the test's sensitivity to actual miscalibration
(Tesso & Vehtari, [2026](#Tesso2026)).

**Deprecation and Compatibility**

As of `bayesplot v1.16.0`, the `"independent"` method is officially superseded.
But it will remain the default method for the time being to simplify the 
transition for users. If no method is specified, `"independent"` will be used
and a message will be printed informing the user:
```
ℹ In the next major release, the default `method` will change to 'correlated'.
• To silence this message, explicitly set `method = 'independent'` or
  `method = 'correlated'`.
```
If `"independent"` is explicitly set, it will trigger a message informing the 
user:
```
"The 'independent' method is superseded by the 'correlated' method."
```
This is intended to encourage a transition to the `"correlated"` method, which
will become the default in a future release.

### `method = "correlated"` (new, recommended)
This method employes one of three dependence-aware uniformity tests 
(selected via the `test` argument; details below) to compute a global p-value 
for the null hypothesis of uniformity. Unlike the independent method, it 
accounts for the correlation among PIT values (Tesso & Vehtari, 
[2026](#Tesso2026)).

Instead of drawing traditional confidence bands, the plot highlights ECDF
regions in red where the pointwise contribution to the test statistic is
largest. This visualization makes it easier to diagnose the *type* and
*location* of miscalibration.

```{r comparison, opts.label="fig_wide"}
set.seed(2026)
pit <- rbeta(300, 1, 1.2)

p1 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "independent"
  ) +
  labs(subtitle = "method = 'independent'")

p2 <- ppc_loo_pit_ecdf(
  pit = pit, method = "correlated"
  ) +
  labs(subtitle = "method = 'correlated'")

p1 | p2
```

## The three uniformity-tests within `method = "correlated"`
When using `method = "correlated"`, you select the testing procedure via the
`test` argument. The three options correspond to the three procedures proposed
by Tesso & Vehtari ([2026](#Tesso2026)):

### `test = "POT"`
Pointwise Order Tests Combination (`POT`) is based on order statistics of the
PIT values and is the **recommended default** for most workflows. It should
ideally use continuous LOO-PITs and works for PSIS-LOO weighted rank-based PITs.

### `test = "PRIT"`
Pointwise Rank-based Individual Tests Combination (`PRIT`) is designed for
**rank-based or discrete PIT values**. It is most compatible with PITs computed
as normalized ranks and is weaker than "POT" in testing weighted rank based
LOO-PIT values for the same reason previously outlined.

### `test = "PIET"`
Pointwise Inverse-CDF Evaluation Tests Combination (`PIET`) applies an 
inverse-CDF transformation before testing and is **tail-sensitive**. It is
exclusively recommended for detecting deviations in tails (or outliers) and
requires continuous PIT values for reliable behavior.

## Additional arguments
With the introduction of the `method = "correlated"` option, the three 
functions now have additional arguments that control the appearance and 
behavior of the plot when using correlated testing procedures. 
These arguments are:

### `gamma`
When `method = "correlated"` and the global test rejects uniformity, the plot
uses color coding to highlight influential regions. The `gamma` argument is a
non-negative numeric threshold: a point is highlighted if its pointwise
contribution to the test statistic exceeds `gamma`. The default is `0`, which
highlights all points that contribute at all when the test rejects. Increasing
`gamma` raises the bar, flagging only the most strongly deviant points.

```{r gamma-example, opts.label="fig_wide"}
set.seed(2026)
pit <- rbeta(300, 1, 1.2)

# Highlight only the most strongly influential regions
p1 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "correlated",
  plot_diff = TRUE,
  gamma = 0
  ) +
    labs(subtitle = "gamma = 0")

# Highlight more broadly
p2 <- ppc_loo_pit_ecdf(
  pit = pit, method = "correlated",
  plot_diff = TRUE,
  gamma = 80
  ) +
  labs(subtitle = "gamma = 80")

p1 | p2
```

### `linewidth` and `color`
`linewidth` controls the width of the ECDF line (default `0.3`). `color` is a
named character vector with two elements controlling the appearance of the plot:
`c(ecdf = "grey60", highlight = "red")`. Whereby `ecdf` is the color of the
main ECDF line; `highlight` is the color used for the
flagged regions when the global test rejects. Both can be overridden:

```{r color-example, opts.label="fig_wide"}
pit <- rbeta(300, 2, 2)

p1 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "correlated"
) +
  labs(subtitle = "default style")

p2 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "correlated",
  color = c(ecdf = "steelblue", highlight = "orange"),
  linewidth = 0.5
) +
  labs(subtitle = "custom color / linewidth")

p1 | p2
```

### `help_text`
Setting `help_text = TRUE` adds an annotation to the plot showing the `test`
name, (1-`prob`) level ($\alpha$), and global p-value. This is the **default**
(`help_text = TRUE`). Set `help_text = FALSE` to suppress the annotation.

```{r help-text-example, opts.label="fig_wide"}
pit <- rbeta(300, 1, 1.2)

p1 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "correlated",
  plot_diff = TRUE,
  help_text = TRUE
) +
  labs(subtitle = "help_text = TRUE")

p2 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "correlated",
  plot_diff = TRUE,
  help_text = FALSE
) +
  labs(subtitle = "help_text = FALSE")

p1 | p2
```

### `help_text_shrinkage`
When `help_text = TRUE`, `help_text_shrinkage` controls the relative size of
the annotation. It is a numeric factor in $(0, 1]$ (default `0.8`) multiplied
by the theme text size. Decrease it to shrink crowded annotations (especially
in faceted / grouped plots); increase it toward `1` for larger labels.

```{r help-text-shrinkage-example, opts.label="fig_wide"}
pit <- rbeta(300, 1, 1.2)

p1 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "correlated",
  plot_diff = TRUE,
  help_text_shrinkage = 0.5
) +
  labs(subtitle = "help_text_shrinkage = 0.5")

p2 <- ppc_loo_pit_ecdf(
  pit = pit,
  method = "correlated",
  plot_diff = TRUE,
  help_text_shrinkage = 1
) +
  labs(subtitle = "help_text_shrinkage = 1")

p1 | p2
```

### `pareto_pit`
The `pareto_pit` argument controls how PIT values are computed internally when
precomputed `pit` values are not supplied. It follows the standard PIT
computation procedure, except that in the tails it substitutes the empirical CDF
(ECDF) with the CDF of a fitted generalized Pareto distribution. This reduces
variability and yields non-zero, non-unity PIT values even outside the range of
the reference draws — making it particularly useful for stabilizing PIT
uniformity checks, where a single 0 or 1 can otherwise introduce large
variation.

The following figure compares simulated PIT values from `rstantools::loo_pit()`
(x-axis) with those from `posterior::pareto_pit()` (y-axis). Both axes are
logit-transformed, so PIT values of exactly 0 or 1 map to `-Inf` or `Inf`,
respectively; these are shown as red bars on the left (for 0) and right (for 1)
edge of the y-axis. The plot illustrates how `pareto_pit` replaces these extreme
values with finite, less extreme alternatives, thereby reducing variability.

```{r pareto_pit-example, opts.label="fig_square"}
set.seed(122)
n <- 300
S <- 100

# True data-generating process
y <- rnorm(n, mean = 0, sd = 1)
yrep <- matrix(rt(n * S, df = 100), nrow = S, ncol = n)

# Log-likelihood: depends only on y (not the draw), so compute once and broadcast
log_lik_vec <- dt(y, df = 100, log = TRUE)
log_lik <- matrix(rep(log_lik_vec, each = S), nrow = S, ncol = n)

# LOO log importance weights via PSIS
psis_result <- loo::psis(-log_lik)
lw <- weights(psis_result, log = TRUE)

# Compute and sort PITs once; reuse sorted vectors for both tail subsets
pit <- rstantools::loo_pit(object = yrep, y = y, lw = lw)
pit2 <- posterior::pareto_pit(x = yrep, y = y, weights = lw, log = TRUE)

pit_sorted <- sort(pit)
pit2_sorted <- sort(pit2)

# Lower- and upper-tail subsets (M points each)
df <- data.frame(pit = pit_sorted, pit2 = pit2_sorted)
df$clipped_0 <- df$pit == 0 | df$pit2 == 0
df$clipped_1 <- df$pit == 1 | df$pit2 == 1

df_normal <- df[!df$clipped_0 & !df$clipped_1, ]
df_clipped_0 <- df[df$clipped_0, ]
df_clipped_1 <- df[df$clipped_1, ]

breaks <- c(0.001, 0.01, 0.1, 0.5, 0.9, 0.99, 0.999)

ggplot(mapping = aes(x = pit, y = pit2)) +
  geom_point(data = df_normal, colour = "black", size = 1) +
  geom_rug(data = df_clipped_0, colour = "red", linewidth = 0.7,
   length = unit(0.03, "npc"), sides = "l") +
  geom_rug(data = df_clipped_1, colour = "red", linewidth = 0.7,
    length = unit(0.03, "npc"), sides = "r") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  scale_x_continuous(transform = "logit", breaks = breaks, labels = breaks) +
  scale_y_continuous(transform = "logit", breaks = breaks, labels = breaks) +
  labs(x = "pit() [logit transform]", y = "pareto_pit() [logit transform]",
    title = "PIT values from pit() vs. pareto_pit()") +
  bayesplot_theme_get()
```

**Default behavior (recommended):** leave `pareto_pit` at its default value.
The function chooses the appropriate PIT computation automatically based on
`method` and `test`:

| method | test | `pareto_pit` default |
|:---|:------:|:------|
| `"independent"` | -- | `FALSE` (empirical/rank-based PIT) |
| `"correlated"` | `"PRIT"` | `FALSE` (empirical/rank-based PIT) |
| `"correlated"` | `"POT"` | `TRUE` (Pareto-smoothed PIT) |
| `"correlated"` | `"PIET"` | `TRUE` (Pareto-smoothed PIT) |

When `pareto_pit = TRUE`, `ppc_loo_pit_ecdf()` requires LOO importance weights
to be supplied (via `lw` or `psis_object`), since the LOO-PIT computation is
weighted. `ppc_pit_ecdf()` and `ppc_pit_ecdf_grouped()` do not require weights;
they call `posterior::pareto_pit()` with `weights = NULL`.

**When to override:** You can set `pareto_pit` manually if you need to force a
specific PIT computation path for diagnostic comparison. In practice this is
rarely necessary.

## Grouped checks with `ppc_pit_ecdf_grouped()`
`ppc_pit_ecdf_grouped()` applies the same PIT-ECDF checks within levels of a
grouping variable (for example a treatment indicator or categorical covariate).
With `method = "correlated"`, the uniformity test, highlighting, and optional
help-text annotation are computed **separately for each group**, and the panels
are arranged with `facet_wrap()`.

This is useful when overall calibration looks acceptable but a subgroup is
miscalibrated. The example below simulates two groups of equal size: group A is
well calibrated, while group B has systematically too-low predictions
(`ncp = -1`).

```{r grouped-example, opts.label="fig_grouped"}
set.seed(2026)
n <- 200
S <- 100
group <- rep(c("A (calibrated)", "B (biased)"), each = n / 2)

y <- rnorm(n, mean = 0, sd = 1)
yrep <- matrix(NA_real_, nrow = S, ncol = n)

idx_a <- group == "A (calibrated)"
idx_b <- !idx_a
yrep[, idx_a] <- matrix(rt(S * sum(idx_a), df = 100, ncp = 0), nrow = S)
yrep[, idx_b] <- matrix(rt(S * sum(idx_b), df = 100, ncp = -1), nrow = S)

ppc_pit_ecdf_grouped(
  y = y,
  yrep = yrep,
  group = group,
  method = "correlated",
  plot_diff = TRUE
)
```

All correlated-method arguments (`test`, `gamma`, `linewidth`, `color`,
`help_text`, `help_text_shrinkage`, `pareto_pit`) work the same way as in
`ppc_pit_ecdf()`. In faceted plots, a smaller `help_text_shrinkage` is often
helpful so annotations do not dominate the panels.

## Using `brms::pp_check()`
It is also possible to use `brms::pp_check()` with `type = "loo_pit_ecdf"` to
perform the same testing and plotting procedure as `ppc_loo_pit_ecdf()`. The
following code snippet provides an example:
```{r pp-check-example, fig.show="hold", out.width="80%", eval=FALSE}
brms::pp_check(
  fit_normal,
  type = "loo_pit_ecdf",
  method = "correlated"
)
```

## Case Studies
Several case studies on the companion website for *Bayesian Workflow*
(Gelman et al., [2026](#Gelman2026)) illustrate how to use the updated 
functions:

+ [Roaches cross-validation model checking and comparison](https://users.aalto.fi/~ave/_site/roaches/roaches.html)
+ [Nabiximols treatment efficiency](https://users.aalto.fi/~ave/_site/nabiximols/nabiximols.html)
+ [Sleep study: Prior specification and model checking](https://users.aalto.fi/~ave/_site/sleep_study/sleep_study.html)
+ [World Cup continuous vs discrete models](https://users.aalto.fi/~ave/_site/world_cup/world_cup.html)

## Summary of new arguments

| Argument | Values | Default |
|---|---|---|
| `method` | `"independent"`, `"correlated"` | `NULL` → falls back to `"independent"` with a deprecation notice |
| `test` | `"POT"`, `"PRIT"`, `"PIET"` | `"POT"` |
| `gamma` | numeric $\ge 0$ | `0` |
| `linewidth` | numeric | `0.3` |
| `color` | named character vector | `c(ecdf = "grey60", highlight = "red")` |
| `help_text` | `TRUE`, `FALSE` | `TRUE` |
| `help_text_shrinkage` | numeric in $(0, 1]$ | `0.8` |
| `pareto_pit` | `TRUE`, `FALSE` | `TRUE` if `test` is `"POT"` or `"PIET"` and no `pit` supplied; `FALSE` otherwise |

## References

Gelman, A., Vehtari, A., McElreath, R., Simpson, D., Margossian, C., Yao, Y., 
Kennedy, L., Gabry, J., Burkner, P.-C., Modrak, M., & Leos Barajas, V. (2026). 
Bayesian Workflow. Routledge. 
https://www.routledge.com/Bayesian-Workflow/Gelman-Vehtari-McElreath-Simpson-Margossian-Yao-Kennedy-Gabry-Burkner-Modrak-Barajas/p/book/9780367490140
<a id="Gelman2026"></a>

Säilynoja, T., Bürkner, P.-C., and Vehtari, A. (2022). Graphical test for
discrete uniformity and its applications in goodness-of-fit evaluation and
multiple sample comparison. *Statistics and Computing*, 32, 32.
https://link.springer.com/10.1007/s11222-022-10090-6
<a id="Säilynoja2022"></a>

Tesso, H., & Vehtari, A. (2026). LOO-PIT predictive model checking.
http://arxiv.org/abs/2603.02928
<a id="Tesso2026"></a>

Vehtari, A., Gelman, A., and Gabry, J. (2017). Practical Bayesian model
evaluation using leave-one-out cross-validation and WAIC. *Statistics and
Computing*, 27(5), 1413–1432.
https://link.springer.com/article/10.1007/S11222-016-9696-4
<a id="Vehtari2017"></a>

Vehtari, A., Simpson, D., Gelman, A., Yao, Y., and Gabry, J. (2024). Pareto
smoothed importance sampling. *Journal of Machine Learning Research*, 25(72),
1–58. https://www.jmlr.org/papers/v25/19-556.html
<a id="Vehtari2024"></a>